---
title: "Kleiner lab database testing"
output: html_notebook
---

Here we are evaluating the preliminary data shared by the Kleiner lab at NCSU to determine if their metaproteomic database should be adjusted (e.g. filtering out homologs to human proteins, other ideas?) before continuing with analysis of the DFC samples.

# R setup

```{r setup, include=FALSE, echo=FALSE}
# This is to make this compatible with version control in git, ok to ignore/exclude
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here) # For relative file paths
library(readxl) # For reading Excel files
library(tidyverse) # For lots of nice "data wrangling" tools in R
```

# Read in files

## Current reference database

```{r}
# Get the path to the file of interest
f <- here('data', 'processed', 'DietaryDB_ForDavidLab.xlsx')
f
```

```{r}
# Read in the first sheet of the file using read_excel function
db <- read_excel(f, sheet = 1)
```

## Participant metaproteomic data

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Cmd+Option+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Cmd+Shift+K* to preview the HTML file). 

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.

