---
title: "Metaproteomic pipeline comparison"
output: html_notebook
---

Abbreviations: 

- PUP: Protein unique peptide
- UP: Unique peptide (*i.e.* unique to the peptide group)

# R setup 

```{r setup, include = FALSE, echo = FALSE}
require(knitr)
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(eulerr) # for Euler diagrams
library(here)
library(phyloseq)
library(scales)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title.x = element_text(size = 14,
                                                face = 'bold'),
                    axis.title.y = element_text(size = 14,
                                                face = 'bold'),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
               )
)
```


# Read in files

## Participant metaproteomic data

This is *not* pre-filtered by Ali.  Read in and do that filtering here.

### Healthy donor

```{r }
# List files
fs <-
     here('data', 
          'raw',
          '2_All5FDRProteins_Excel_Files_DB2.1Search_Healthy donor') %>%
     list.files(pattern = ".xlsx", full.names = T)

# Read and synchronize names
# Note that if sheet not specified, defaults to first sheet
dfs <- 
     lapply(fs, read_excel) %>% 
     # Rename "Found in sample" column so it's the same for each sample
     lapply(., 
            function(x){
                 # Update name
                 names(x)[grep('Found in Sample: *', names(x))] <- 'Found in Sample'
                 # Return original object
                 x
                 }
            )

# Bind together
names(dfs) <- basename(fs)
healthydonor <- bind_rows(dfs, .id = 'id')
rm(fs, dfs)
```

### Weight Loss

```{r }
# List files
fs <-
     here('data', 
          'raw',
          '2_All5FDRProteins_Excel_Files_DB2.1Search_Including Outlier') %>%
     list.files(pattern = ".xlsx", full.names = T)

# Read and synchronize names
# Note that if sheet not specified, defaults to first sheet
dfs <- 
     lapply(fs, read_excel) %>% 
     # Rename "Found in sample" column so it's the same for each sample
     lapply(., 
            function(x){
                 # Update name
                 names(x)[grep('Found in Sample: *', names(x))] <- 'Found in Sample'
                 # Return original object
                 x
                 }
            )

# Bind together
names(dfs) <- basename(fs)
weightloss <- bind_rows(dfs, .id = 'id')
rm(fs, dfs)
```

## phyloseq object

### 1 PUP

Think I can hack this together from two older versions, but probably want to go back and check just to be sure. From README notes:
20210408_ps_mp: v2 of analysis on DFC samples
20211102_ps_mp: v2 of analysis on 908 samples

```{r}
ps.pup.dfc <- 
     here('data', 
          'processed', 
          'phyloseq',
          '20210408_ps_mp.rds') %>% 
     readRDS()

ps.pup.dfc
```

```{r}
ps.pup.908 <- 
     here('data', 
          'processed', 
          'phyloseq',
          '20211102_ps_mp.rds') %>% 
     readRDS()

ps.pup.908
```

```{r}
ps.pup <- merge_phyloseq(ps.pup.dfc, ps.pup.908)
ps.pup
```

### 1 UP

```{r}
ps.up <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'ps_mp.rds') %>% 
     readRDS()

ps.up
```

# Pre-process

## Participant metaproteomic data

### Combine

```{r}
# Clean up subject IDs
head(weightloss$id)
head(healthydonor$id)
```

```{r}
weightloss <- 
     weightloss %>% 
     mutate(id = gsub(id,
                      pattern = '^\\d+_',
                      replacement = '')) %>% 
     separate(id,
              into = c('subj', 'date'),
              sep = '_', 
              extra = 'drop') %>% 
     mutate(date = as.Date(date, 
                           format = '%d%b%y'))
```


```{r}
healthydonor <- 
     healthydonor %>% 
     # Clean up ID column
     mutate(subj = gsub(id,
                        pattern = '_.*$',
                        replacement = '')
            ) %>% 
     # Convert date to Date type
     mutate(date = 
                 case_when(
                      subj == 'HH1' ~ as.Date('25Jun18', 
                                              format = '%d%b%y'),
                      subj == 'HH2' ~ as.Date('26Jun18', 
                                              format = '%d%b%y'),
                      subj == 'HH3' ~ as.Date('27Jun18', 
                                              format = '%d%b%y'),
                      subj == 'HH4' ~ as.Date('28Jun18', 
                                              format = '%d%b%y'),
                      subj == 'HH5' ~ as.Date('29Jun18', 
                                              format = '%d%b%y'))
     ) %>% 
     # Can now overwrite subject
     mutate(subj = '908') 
```


```{r}
# Add study
weightloss$study <- 'Weight Loss'
healthydonor$study <- 'Healthy Donor'
```

```{r}
# Combine
proteins.df <- bind_rows(weightloss, healthydonor)
dim(proteins.df)
rm(weightloss, healthydonor)
```

### Filter

```{r}
# Filter to only food-derived proteins
proteins.df <- 
     filter(proteins.df, 
            !grepl(Accession,
                   pattern = 'Human_|Microbiota_')) 

dim(proteins.df)
```

#### 1PUP

```{r}
range(proteins.df$`# Protein Unique Peptides`, na.rm = TRUE)
```

```{r}
pup.df <- filter(proteins.df, `# Protein Unique Peptides` > 0)
range(pup.df$`# Protein Unique Peptides`, na.rm = TRUE)
```

#### 1UP

```{r}
range(proteins.df$`# Unique Peptides`, na.rm = TRUE)
```

```{r}
up.df <- filter(proteins.df, `# Unique Peptides` > 0)
range(up.df$`# Unique Peptides`, na.rm = TRUE)
```

```{r }
# Quick QC
table(pup.df$`Found in Sample`)
table(up.df$`Found in Sample`)
```

```{r}
# Hmm: what are these 6?
filter(pup.df, `Found in Sample` == 'Not Found')
```
Interestingly these are all 'None' in the Master column.  

```{r}
# Remove
pup.df <- filter(pup.df, `Found in Sample` != 'Not Found')
```

### Master proteins

```{r}
# For unique peptide analysis only: Filter to master proteins
table(pup.df$Master)
```

```{r}
table(up.df$Master)
```

Okay, so in the UP analysis we want to consider master proteins only.  Not sure what to do about the stragglers in the 1PUP dataset but they are few enough that I think safe to exclude them also.  Some look suspiciously like dietary proteins:

```{r}
pup.df %>% filter(Master != 'Master Protein') %>% pull(Description)
```

```{r}
# Filter
pup.df <- filter(pup.df, Master == 'Master Protein')
up.df <- filter(up.df, Master == 'Master Protein')
```

### Combine
```{r}
# For joint plotting
up.df$analysis <- '1UP'
pup.df$analysis <- '1PUP'

proteins.df <-
     bind_rows(up.df,
               pup.df)
```

```{r}
# Check completeness of data
# A few hundred missing entries for #PUPs when >1UP
filter(proteins.df, is.na(`# Protein Unique Peptides`))

# Replace these with 0
proteins.df$`# Protein Unique Peptides`[is.na(proteins.df$`# Protein Unique Peptides`)] <-  0
```


# Analyze

## Compare raw datasets

```{r}
# How many additional PSMs?
proteins.df %>% 
     group_by(analysis) %>% 
     summarize(psms_tot = sum(`# PSMs`))
```
```{r}
95289-68766
```

```{r}
95289/68755
```

```{r}
# Visualize PSM counts
proteins.df %>% 
     filter(analysis == '1UP') %>% 
     dplyr::rename(pups = `# Protein Unique Peptides`) %>% 
     group_by(study, pups > 0) %>% 
     summarize(psms_tot = sum(`# PSMs`)) %>% 
     ggplot(aes(x = study,
                y = psms_tot,
                fill = `pups > 0`)) +
     geom_col(position = 'stack') +
     scale_y_continuous(label = comma) +
     scale_fill_manual(values = c('#deac13', '#f1ce63')) +
     labs(y = '# PSMs') +
     theme(panel.border = element_blank(),
           axis.line = element_line(),
           panel.grid = element_blank(),
           axis.title.x = element_blank(),
           legend.position = 'none')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            'filtering',
            paste0(plotID, '_# PSMs by analysis version.pdf')),
       height = 3, width = 3)
```

```{r}
# What percentage of total are from master proteins without PUPs?
up.df %>% 
     filter(`# Protein Unique Peptides` == 0) %>% 
     pull(`# PSMs`) %>% 
     sum()
```
```{r}
26118/sum(up.df$`# PSMs`)
```

```{r}
# Visualize protein counts
proteins.df %>% 
     filter(analysis == '1UP') %>% 
     dplyr::rename(pups = `# Protein Unique Peptides`) %>% 
     group_by(study, pups > 0) %>% 
     count() %>% 
     ggplot(aes(x = study, 
                y = n, 
                fill = `pups > 0`)) +
     geom_bar(stat = 'identity', position = 'stack') +
     scale_fill_manual(values = c('#deac13', '#f1ce63')) +
     scale_y_continuous(label = comma) +
     labs(y = '# proteins') +
     # ylim(0, 60000) +
     theme(panel.border = element_blank(),
           panel.grid = element_blank(),
           axis.title.x = element_blank(),
           axis.line = element_line(),
           legend.position = 'none')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'supplemental',
            'filtering',
            paste0(plotID, '_# proteins by analysis version.pdf')),
       height = 3, width = 3)
```

```{r}
# How many additional proteins?
proteins.df %>% 
     group_by(analysis) %>% 
     summarize(proteins_tot = n_distinct(Accession))
```

```{r}
8398-5782	
```


```{r}
# What percentage of total are from master proteins without PUPs?
up.df %>% 
     filter(`# Protein Unique Peptides` == 0) %>% 
     dim()
```
```{r}
4238/dim(up.df)[1]
```

## Compare taxa

What do we detect with "relaxed" analysis that we don't see if we require at least one protein unique peptide?

```{r}
setdiff(taxa_names(ps.up), taxa_names(ps.pup)) %>% sort()
```

And reverse?

```{r}
setdiff(taxa_names(ps.pup), taxa_names(ps.up))
```

```{r}
# Amaranth is a difference in NCBI taxonomy between dates, rename
taxa_names(ps.pup)[taxa_names(ps.pup) == 'Amaranthus hybridus subsp. cruentus'] <- 'Amaranthus cruentus'
```


### Euler plot

```{r}
# Get union of taxa detected in the two analyses
all <- 
     union(taxa_names(ps.pup),
           taxa_names(ps.up))

length(all)
```

```{r}
# Make dataframe for plot input
plot.df <- 
     data.frame(taxon = all) %>% 
     mutate(up = taxon %in% taxa_names(ps.up),
            pup = taxon %in% taxa_names(ps.pup))

# Quick checks
sum(plot.df$up) 
sum(plot.df$pup) 
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
pdf(here('results', 
         'manuscript',
         'supplemental',
         'filtering',
         paste0(plotID, '_Taxa Euler diagram (metaproteomic analysis), with counts.pdf')),
    height = 3, width = 3)

euler(plot.df[,c('up', 'pup')]) %>% 
     plot(fills = c('#deac13', '#f1ce63'),
          labels = FALSE,
          shape = "ellipse", 
          quantities = TRUE)
          # quantities = FALSE)
```

