---
title: "Protein percentage breakdown"
output: html_notebook
---

This notebook processe the unfiltered metaproteomic data (which includes dietary, host, and microbial peptides) to determine the percent of proteinaceous biomass from each

# R setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, echo=FALSE}
library(here) 
library(phyloseq) 
library(readxl) # For parsing Excel files
library(tidyverse) 
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title.x = ggtext::element_markdown(size = 14,
                                                            face ='bold'),
                    axis.title.y = ggtext::element_markdown(size = 14,
                                                            face ='bold'),
                    legend.title = element_text(size = 12,
                                                face ='bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
               )
)
```

```{r}
study.color <- c('#4e79a7', # Weight Loss
                 '#f28e2b') # Healthy Donor
```

# Read in files

## Healthy donor

```{r }
# List files
fs <-
     here('data', 
          'raw',
          '2_All5FDRProteins_Excel_Files_DB2.1Search_Healthy donor') %>%
     list.files(pattern = ".xlsx", full.names = T)

# Read and synchronize names
# Note that if sheet not specified, defaults to first sheet
dfs <- 
     lapply(fs, read_excel) %>% 
     # Rename "Found in sample" column so it's the same for each sample
     lapply(., 
            function(x){
                 # Update name
                 names(x)[grep('Found in Sample: *', 
                               names(x))] <- 'Found in Sample'
                 # Return original object
                 x
                 }
            )

# Bind together
names(dfs) <- basename(fs)
proteins.df <- bind_rows(dfs, .id = 'id')
rm(fs, dfs)
```

## Weight loss

```{r }
# List files
fs <-
     here('data', 
          'raw',
          '2_All5FDRProteins_Excel_Files_DB2.1Search_Including Outlier') %>%
     list.files(pattern = ".xlsx", full.names = T)

# Read and synchronize names
# Note that if sheet not specified, defaults to first sheet
dfs <- 
     lapply(fs, read_excel) %>% 
     # Rename "Found in sample" column so it's the same for each sample
     lapply(., 
            function(x){
                 # Update name
                 names(x)[grep('Found in Sample: *', 
                               names(x))] <- 'Found in Sample'
                 # Return original object
                 x
                 }
            )

# Bind together
names(dfs) <- basename(fs)
proteins.df <- 
     bind_rows(dfs, .id = 'id') %>% 
     bind_rows(proteins.df) # Bind back to 908 dataframe created above

rm(fs, dfs)
```

# Pre-process

## Simplify names

```{r}
unique(proteins.df$id)
```

```{r}
proteins.df <- 
     proteins.df %>% 
     mutate(
          # Healthy donor pattern
          id = gsub(id,
                    pattern = '_RealisticFoodsDB.*',
                    replacement = ''),
          # Weight loss pattern
          id = gsub(id,
                    pattern = '_AB\\d+_.*',
                    replacement = ''),
          id = gsub(id,
                    pattern = '^\\d+_',
                    replacement = ''))
```

```{r}
# Synchronize date format
proteins.df$id[proteins.df$id == 'HH1'] <- '908_25Jun18'
proteins.df$id[proteins.df$id == 'HH2'] <- '908_26Jun18'
proteins.df$id[proteins.df$id == 'HH3'] <- '908_27Jun18'
proteins.df$id[proteins.df$id == 'HH4'] <- '908_28Jun18'
proteins.df$id[proteins.df$id == 'HH5'] <- '908_29Jun18'
```

```{r}
# Separate and format as date, then recombine
proteins.df <- 
     proteins.df %>% 
     separate(id, 
              into = c('subj', 'date'),
              sep = '_') %>% 
     mutate(date = as.Date(date,
                           format = '%d%b%y')) %>% 
     mutate(id = paste(subj, date, sep = '_')) %>% 
     select(id, everything())
```

```{r}
n_distinct(proteins.df$id) 
```

```{r}
# Add study variable for downstream faceting
proteins.df <- 
     mutate(proteins.df,
            study = ifelse(subj == '908',
                           yes = 'Healthy donor',
                           no = 'Weight loss'))
```

## Remove outlier

Outlier sample is 15_ML48_17Oct19_AB006. 

```{r}
proteins.df <- 
     filter(proteins.df,
            id != 'ML48_2019-10-17')     
```

```{r}
n_distinct(proteins.df$id) 
```

```{r}
str(proteins.df)
```

## Filter master proteins

```{r}
table(proteins.df$Master)
```

```{r}
# PSMs before filter
sum(proteins.df$`# PSMs`)
```

```{r}
proteins.df <- 
     filter(proteins.df, 
            Master == 'Master Protein')
```

```{r}
# PSMs after filter
sum(proteins.df$`# PSMs`)
```

```{r}
839318/2324012
```

TODO: Go through this logic again to make sure I am understanding.  Does this filtering change regardless of whether I'm using PUP or UP analysis?

## QC parameters

```{r}
table(proteins.df$Checked)
```

```{r}
table(proteins.df$`Protein FDR Confidence: Combined`)
```

## Format accession column

Pull this information from the accession column, which gives a prefix for the taxon:

```{r}
head(proteins.df$Accession)
```

```{r}
proteins.df <- 
     proteins.df %>% 
     mutate(category = gsub(Accession,
                            pattern = '_.*$',
                            replacement = ''))

n_distinct(proteins.df$category)
```

There are general categories ("Human" and "Microbiota"), and then all foods would map to a specific food taxon name. So want to collapse these.

```{r}
proteins.df %>% 
     pull(category) %>% 
     unique() %>% 
     head()
```

Hmmm, it looks like not all have the prefix. However, believe these are all foods. Check for human or microbiota in OS entries:

```{r}
os <- 
     proteins.df %>% 
     filter(!grepl(Accession,
                   pattern = '_')) %>% 
     pull(Description)

head(os)
```

```{r}
os %>% 
     gsub(pattern = '^.+OS=',
          replacement = '') %>% 
     gsub(pattern = ' OX=.+',
          replacement = '') %>% 
     unique() %>% 
     sort()
```

Okay, all foods.  Now can relabel with simplified categories
```{r}
proteins.df <- 
     proteins.df %>% 
     mutate(category = ifelse(!(category %in% c('Human',
                                                'Microbiota')),
                              yes = 'Diet',
                              no = category))
```

```{r}
unique(proteins.df$category)
```

# Analyze

## Overall dataset

```{r}
proteins.df %>% 
     group_by(category) %>% 
     summarize(psms = sum(`# PSMs`)) %>% 
     mutate(psms_pct = psms/sum(psms))
```

## Per-sample

So here, per sample, want to sum PSMs for each category.  

```{r}
psms <- 
     proteins.df %>% 
     group_by(study, id, subj, date, category) %>% 
     summarize(psms = sum(`# PSMs`))
```

## Visualize

```{r}
ggplot(psms, aes(x = id, y = psms, fill = category)) +
     geom_col(position = 'stack') +
     facet_wrap(~study + subj,
                scales = 'free_x') +
     theme(axis.text.x = element_blank())
```
## Percentages

By study (long gradient for 908 vs shorter gradient for DFC samples):

```{r}
ggplot(psms, aes(x = id, y = psms, fill = category)) +
     geom_col(position = 'fill') +
     labs(y = 'Proteinaceous<br>biomass (%)') +
     scale_fill_manual(values = c('#f1ce63',
                                  '#a3acb9',
                                  '#7b848f')) +
     theme(axis.ticks.x = element_blank(),
           axis.text.x = element_blank(),
           legend.title = element_blank(), 
           panel.border = element_blank(), 
           panel.grid = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            'description',
            paste0(plotID, '_Proteinaceous biomass.pdf')),
       height = 3, width = 7)
```


By subject: 
```{r}
proteins.df %>% 
     group_by(subj, date, category) %>% 
     summarize(psms = sum(`# PSMs`)) %>% 
     group_by(subj, date) %>% 
     mutate(psms_tot = sum(psms),
            psms_pct = psms/psms_tot) %>% 
     ungroup() %>% 
     select(-c(psms, psms_tot)) %>% 
     pivot_wider(names_from = category,
                 values_from = psms_pct) %>% 
     group_by(subj) %>% 
     summarize(across(.cols = Diet:Microbiota,
                      .fns = list(mean = mean, sd = sd)))
```
