---
title: "Specific food plots"
output: html_notebook
---

# R setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, echo=FALSE}
library(cowplot)
library(here) 
library(phyloseq) 
library(tidyverse) 
```

```{r}
colors.detection <- 
     c('#849db1', 
       '#fbb04e',
       '#7e756d')

names(colors.detection) <- 
     c('DNA',
       'Protein',
       'Menu')
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title.x = ggtext::element_markdown(size = 14),
                    axis.title.y = ggtext::element_markdown(size = 14),
                    legend.title = element_text(size = 12),
                    strip.text = element_text(size = 12)
               )
)
```

# Read in data

## Taxon name mapping

```{r}
name.sync <- 
     here('data',
          'processed',
          'Naming discrepancies.xlsx') %>% 
     readxl::read_excel()

name.sync
```
## Sample dates

```{r}
samples <- 
     here('data',
          'metadata',
          'Sample collection times.csv') %>% 
     read_csv()

samples
```

## Metaproteomic phyloseq

```{r}
ps.protein <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          # 'ps_mp_1PUP.rds') %>% # >=1 protein unique peptide
          'ps_mp_1UP.rds') %>% # >=1 unique peptide
     readRDS()
          
ps.protein
```

## Menu data phyloseq

```{r}
ps.menu <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_menu.rds') %>% 
     readRDS()

ps.menu
```

### Protein annotations

```{r}
# Raw protein list
proteins.df <- 
     here('data',
          'processed',
          '5FDR_1UP_both datasets.rds') %>% 
     readRDS()
```

```{r}
annotations <- 
     here('data',
          'processed',
          'Animal-based proteins (full annotation).csv') %>% 
     read_csv()

annotations
```
## Metabarcoding phyloseq

```{r}
ps.mb.plant <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_trnL.rds') %>% 
     readRDS()

ps.mb.plant
```

```{r}
ps.mb.animal <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_12SV5.rds') %>% 
     readRDS()

ps.mb.plant
```

# Pre-process

## Samples

```{r}
samples$date <- as.Date(samples$date,
                        format = '%m/%d/%y')
```

```{r}
# Add detection for each so shading is done by facet, below
detection <- c('Menu', 'DNA', 'Protein')
subj <- unique(samples$subj)

samples <-
     expand.grid(detection = detection, 
                 subj = subj) %>% 
     right_join(samples) %>% 
     arrange(study, subj, date) %>% 
     select(subj:time, detection)

samples
```
## Metaproteomics

### Label proteins

These are already filtered:

```{r}
head(taxa_names(ps.protein))
ps.protein
```

```{r}
annotations <- select(annotations, 
                      protein = word, category)
```

```{r}
cat('Starting entries:', nrow(proteins.df), '\n')

proteins.df <- 
     proteins.df %>% 
     select(-category) %>% # 'plant' or 'animal'; want to replace
     filter(Accession %in% taxa_names(ps.protein)) %>% 
     mutate(protein = gsub(Description,
                           pattern = ' OS=.*$',
                           replacement = '')) %>% 
     left_join(annotations)

cat('Filtered entries:', nrow(proteins.df))
```

```{r}
n_distinct(proteins.df$Accession)
```

Great! This lines up with the number of taxa in the phyloseq object. Subset to relevant columns and use for filtering later:

```{r}
annotations <- 
     proteins.df %>% 
     select(Accession, category, protein) %>% 
     distinct() %>% 
     mutate(category = ifelse(is.na(category),
                              yes = 'other',
                              no = category))

rm(proteins.df)
```

```{r}
table(annotations$category)
```

## Menu

Nothing to do here; all handled in combined phyloseq notebooks.
Is this all menu days? Yes:

```{r}
ps.menu
```


## Subset and melt

Currently I'm concerned only with the following examples of tissue-type biomarkers:
- Chicken and egg
- Cattle and dairy

```{r}
# Filter objects to only these taxa
ps.protein <- 
     subset_taxa(ps.protein, 
                 species %in% c('Bos taurus',
                                'Gallus gallus'))

ps.menu <- 
     subset_taxa(ps.menu, 
                 species %in% c('Bos taurus',
                                'Gallus gallus'))
```

```{r}
melt.menu <- psmelt(ps.menu) 
melt.mp <- psmelt(ps.protein)
```

```{r}
# Add common name
melt.menu <- 
     melt.menu %>% 
     left_join(select(name.sync,
                      menu_taxa, 
                      mp_to_menu),
               by = c('name' = 'menu_taxa'))

melt.mp <- 
     melt.mp %>% 
     left_join(select(name.sync,
                      mp_taxa, 
                      mp_to_menu),
               by = c('name' = 'mp_taxa'))
```

```{r}
# Add dataset info to each melted object
melt.menu$detection <- 'Menu'
melt.mp$detection <- 'Protein'
```

```{r}
# Add tissue type
# Menu: already done
```

```{r}
# Metaproteomics
# Sum abundances within each food taxon and tissue type
melt.mp <- 
     left_join(melt.mp, annotations,
               by = c('OTU' = 'Accession'))

melt.mp <- 
     melt.mp %>% 
     group_by(detection, 
              study,
              subj, 
              date,
              mp_to_menu, 
              food_group = category) %>% 
     summarize(Abundance = sum(Abundance)) %>% 
     ungroup()
```

#### Join melts

```{r}
melts <- list(melt.menu, melt.mp)
```

```{r}
vs <- c('study',
        'subj', 
        'date',
        'detection', 
        'mp_to_menu', 
        'food_group',
        'Abundance') # Shared variables to pull from each phyloseq

lapply(melts, ncol)
melts <- lapply(melts, function(x){select(x, one_of(vs))})
lapply(melts, ncol)
```

```{r}
melt <- bind_rows(melts)
dim(melt)
rm(melts)
```

```{r}
melt$detection <- factor(melt$detection, 
                         levels = c('Menu',
                                    'Protein'))
```

```{r}
# Synchronize food group naming:
melt %>% 
     group_by(detection) %>% 
     summarize(unique(food_group))
```
```{r}
melt$food_group[melt$food_group == 'muscle'] <- 'meat'
melt$food_group[melt$food_group == 'eggs, shell eggs'] <- 'egg'
melt$food_group[melt$food_group == 'poultry'] <- 'meat'
melt$food_group[melt$food_group == 'dairy, solid/semi-solid dairy products'] <- 'dairy'
melt$food_group[melt$food_group == 'dairy, fluid milk'] <- 'dairy'
```

# Plot

### "Overall" summaries

Note that this is currently with all menu days included.
```{r}
melt %>% 
     filter(mp_to_menu == 'chicken',
            study == 'Intervention') %>% 
     pivot_wider(names_from = detection,
                 values_from = Abundance) %>% 
     ggplot(aes(x = Menu, y = Protein, color = subj)) +
     geom_point(alpha = 0.8) +
     facet_wrap(~food_group)     
```


### Pairing to menu

```{r}
taxa <- c('chicken', 'cattle')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

for (taxon in taxa){
     melt.filt <- filter(melt, common_name == taxon)
     
     ggplot(melt.filt, aes(x = date)) +
          geom_tile(data = samples, aes(x = date, 
                                        y = 0),
                    fill = 'gray90',
                    height = Inf) +
          geom_bar(aes(y = Abundance, fill = detection), stat = 'identity') +
          facet_grid(cols = vars(subj),
                     rows = vars(detection),
                     scales = 'free',
                     space = 'free_x') +
          scale_fill_manual(values = colors.detection) +
          labs(x = 'Date', y = 'Food abundance') +
          theme(axis.text.x = element_blank(),
                legend.position = 'none',
                panel.grid = element_blank())
     
     # ggsave(here('results', 
     #             'biomarkers', 
     #             paste0(plotID, 
     #                    '_',
     #                    taxon,
     #                    ' detection by measure and sample.pdf')),
     #        height = 4)
}
```

### Protein only

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

samples.filt <- filter(samples, detection != 'DNA')

for (taxon in taxa){
     melt.filt <- 
          filter(melt, common_name == taxon & detection != 'DNA')
     
     ggplot(melt.filt, aes(x = date)) +
          geom_tile(data = samples.filt, aes(x = date, 
                                             y = 0),
                    fill = 'gray90',
                    height = Inf) +
          geom_bar(aes(y = Abundance, fill = detection), stat = 'identity') +
          facet_grid(cols = vars(subj),
                     rows = vars(detection),
                     scales = 'free',
                     space = 'free_x') +
          scale_fill_manual(values = colors.detection) +
          labs(x = 'Date', y = 'Food abundance') +
          theme(axis.text.x = element_blank(),
                legend.position = 'none',
                panel.grid = element_blank())
     
     # ggsave(here('results', 
     #             'biomarkers', 
     #             paste0(plotID, 
     #                    '_',
     #                    taxon,
     #                    ' detection by measure and sample.pdf')),
     #        height = 3)
}
```

### Protein by type

#### Cattle 

```{r}
taxon <- 'cattle'
samples.filt <- filter(samples, detection != 'DNA')
melt.filt <- 
     filter(melt, common_name == taxon & detection != 'DNA')
```

##### Recode tissue entries

```{r}
melt.filt %>% 
     group_by(detection) %>% 
     summarize(unique(group))
```

Preserve differences between dairy, but equate meat and muscle.

```{r}
melt.filt$group[melt.filt$group == 'muscle'] <- 'meat'
```

##### Plot

```{r}
colors.cattle <- c(
     '#4e79a7',
     '#4e79a7',
     '#a0cbe8', 
     '#e15759',
     '#bab0ac' # other
                   )
```

Try alternate way of plotting: 1 per subject

```{r}
subjects <- 
     samples %>% 
     pull(subj) %>% 
     unique() %>% 
     sort()

plot.list <- list()

for (s in 1:5){
     if (s != 5){ # Label facet differently in last subject
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.cattle) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     strip.text.y = element_blank(),
                     legend.position = 'none')
     } else {
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         width = 1,
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.cattle) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     legend.position = 'none')
     }
     # Add to plot list
     plot.list[[s]] <- p
}
```

```{r}
# Relative widths:
widths <- 
     melt.filt %>%
     group_by(subj) %>% 
     summarize(dates = n_distinct(date)) %>% 
     ungroup() %>% 
     mutate(relative = dates/sum(dates))

widths
```

```{r}
plot_grid(plotlist = plot.list,
          nrow = 1,
          rel_widths = c(1.2, 2, 2, 2, 2.4))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'biomarkers',
            paste0(plotID,
                   '_',
                   taxon,
                   ' cattle tissue types by sample and menu.pdf')),
       height = 2.5, width = 10)
```

#### Chicken
```{r}
taxon <- 'chicken'
samples.filt <- filter(samples, detection != 'DNA')
melt.filt <- 
     filter(melt, common_name == taxon & detection != 'DNA')
```

##### Recode tissue entries

```{r}
melt %>% 
     group_by(detection) %>% 
     summarize(unique(food_group))
```

```{r}
melt.filt$group[melt.filt$group == 'muscle'] <- 'meat'
melt.filt$group[melt.filt$group == 'poultry'] <- 'meat'
melt.filt$group[melt.filt$group == 'eggs, shell eggs'] <- 'egg'
```

```{r}
colors.chicken <- c('#f1ce63',
                   '#ff9d9a',
                   '#bab0ac')
```

```{r}
subjects <- 
     samples %>% 
     pull(subj) %>% 
     unique() %>% 
     sort()

plot.list <- list()

for (s in 1:5){
     if (s != 5){ # Label facet differently in last subject
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.chicken) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     strip.text.y = element_blank(),
                     legend.position = 'none')
     } else {
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         width = 1,
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.chicken) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     legend.position = 'none')
     }
     # Add to plot list
     plot.list[[s]] <- p
}
```

```{r}
# Relative widths:
widths <- 
     melt.filt %>%
     group_by(subj) %>% 
     summarize(dates = n_distinct(date)) %>% 
     ungroup() %>% 
     mutate(relative = dates/sum(dates))

widths
```

```{r}
plot_grid(plotlist = plot.list,
          nrow = 1,
          rel_widths = c(1.2, 2, 2, 2, 2.4))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'biomarkers',
            paste0(plotID,
                   '_',
                   taxon,
                   ' chicken tissue types by sample and menu.pdf')),
       height = 2.5, width = 10)
```