---
title: "Specific food plots"
output: html_notebook
---

# R setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, echo=FALSE}
library(cowplot)
library(here) 
library(phyloseq) 
library(tidyverse) 
```

```{r}
colors.detection <- 
     c('#849db1', 
       '#fbb04e',
       '#7e756d')

names(colors.detection) <- 
     c('DNA',
       'Protein',
       'Menu')

# Define color palettes
colors.subj <- 
     palette.colors(n = 5, 
                    palette = 'Okabe-Ito') %>% 
     unname()
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title.x = ggtext::element_markdown(size = 14),
                    axis.title.y = ggtext::element_markdown(size = 14),
                    legend.title = element_text(size = 12),
                    strip.text = element_text(size = 12)
               )
)
```

# Read in data

## Taxon name mapping

```{r}
name.sync <- 
     here('data',
          'processed',
          'Naming discrepancies.xlsx') %>% 
     readxl::read_excel()

name.sync
```
## Sample dates

```{r}
samples <- 
     here('data',
          'metadata',
          'Sample collection times.csv') %>% 
     read_csv()

samples
```

## Metaproteomic phyloseq

```{r}
ps.protein <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          # 'ps_mp_1PUP.rds') %>% # >=1 protein unique peptide
          'ps_mp_1UP.rds') %>% # >=1 unique peptide
     readRDS()
          
ps.protein
```

## Menu data phyloseq

```{r}
ps.menu <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_menu.rds') %>% 
     readRDS()

ps.menu
```

### Protein annotations

```{r}
# Raw protein list
proteins.df <- 
     here('data',
          'processed',
          '5FDR_1UP_both datasets.rds') %>% 
     readRDS()
```

```{r}
annotations <- 
     here('data',
          'processed',
          'Animal-based proteins (full annotation).csv') %>% 
     read_csv()

annotations
```

```{r}
table(annotations$category)
```

# Pre-process

## Samples

```{r}
samples$date <- as.Date(samples$date,
                        format = '%m/%d/%y')
```

```{r}
# Add detection for each so shading is done by facet, below
detection <- c('Menu', 'DNA', 'Protein')
subj <- unique(samples$subj)

samples <-
     expand.grid(detection = detection, 
                 subj = subj) %>% 
     right_join(samples) %>% 
     arrange(study, subj, date) %>% 
     select(subj:time, detection)

samples
```
## Metaproteomics

### Label proteins

These are already filtered:

```{r}
head(taxa_names(ps.protein))
ps.protein
```

```{r}
annotations <- select(annotations, 
                      protein = word, category)
```

```{r}
cat('Starting entries:', nrow(proteins.df), '\n')

proteins.df <- 
     proteins.df %>% 
     select(-category) %>% # 'plant' or 'animal'; want to replace
     filter(Accession %in% taxa_names(ps.protein)) %>% 
     mutate(protein = gsub(Description,
                           pattern = ' OS=.*$',
                           replacement = '')) %>% 
     left_join(annotations)

cat('Filtered entries:', nrow(proteins.df))
```

```{r}
n_distinct(proteins.df$Accession)
```

Great! This lines up with the number of taxa in the phyloseq object. Subset to relevant columns and use for filtering later:

```{r}
annotations <- 
     proteins.df |> 
     select(Accession, scientific_name, category, protein) %>% 
     distinct() |> 
     mutate(category = ifelse(is.na(category),
                              yes = 'other',
                              no = category))

rm(proteins.df)
```

```{r}
cat('All organisms:')
table(annotations$category)

cat('\nCattle and chicken only:')
annotations |> 
     filter(scientific_name %in% c('Bos taurus',
                                   'Gallus gallus')) |> 
     pull(category) |> 
     table()
```

## Select menu days

### 1 or 2 days prior

```{r}
nsamples(ps.menu)
nsamples(ps.protein)
```

### 1-2 days prior (matched)

```{r}
# Can't use merge_samples on phyloseq here because some samples make up >1 merge

# Do manually on ASV table
asvtab.menu <- otu_table(ps.menu)@.Data
asvtab.menu.12 <- data.frame()
samdf.protein <- 
     ps.protein@sam_data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row')

for (sample in seq(dim(samdf.protein)[1])){
     # Pull menu day -1
     one.before <- asvtab.menu[samdf.protein$delta1[sample], ]
     # Pull menu day -2
     two.before <- asvtab.menu[samdf.protein$delta2[sample], ]
     # Merge and place in new, aggregated OTU table
     asvtab.menu.12 <- rbind(asvtab.menu.12,
                             (one.before + two.before))
     # Update food names (only has to be done once)
     if (sample == 1){names(asvtab.menu.12) <- 
          names(one.before + two.before)}
     # Update sample name
     row.names(asvtab.menu.12)[sample] <- samdf.protein$row[sample]
}

# Now rebuild a subsetted phyloseq object
ps.menu <- phyloseq(otu_table(asvtab.menu.12, taxa_are_rows = FALSE),
                       sample_data(ps.protein), # Now this matches
                       tax_table(ps.menu)) # Menu-specific taxonomy

# Remove any taxa that aren't present any longer
ps.menu <- prune_taxa(taxa_sums(ps.menu) > 0, ps.menu)
ps.menu
```
## Subset and melt

Currently I'm concerned only with the following examples of tissue-type biomarkers:
- Chicken and egg
- Cattle and dairy

```{r}
# Filter objects to only these taxa
ps.protein <- 
     subset_taxa(ps.protein, 
                 species %in% c('Bos taurus',
                                'Gallus gallus'))

ps.menu <- 
     subset_taxa(ps.menu, 
                 species %in% c('Bos taurus',
                                'Gallus gallus'))
```

```{r}
melt.menu <- psmelt(ps.menu) 
melt.mp <- psmelt(ps.protein)
```

```{r}
# Add common name
melt.menu <- 
     melt.menu %>% 
     left_join(select(name.sync,
                      menu_taxa, 
                      mp_to_menu),
               by = c('name' = 'menu_taxa'))

melt.mp <- 
     melt.mp %>% 
     left_join(select(name.sync,
                      mp_taxa, 
                      mp_to_menu),
               by = c('name' = 'mp_taxa'))
```

```{r}
# Add dataset info to each melted object
melt.menu$detection <- 'Menu'
melt.mp$detection <- 'Protein'
```

```{r}
# Add tissue type
# Menu: already done
```

```{r}
# Metaproteomics
# Sum abundances within each food taxon and tissue type
melt.mp <- 
     left_join(melt.mp, annotations,
               by = c('OTU' = 'Accession'))

melt.mp <- 
     melt.mp %>% 
     group_by(detection, 
              study,
              subj, 
              date,
              mp_to_menu, 
              food_group = category) %>% 
     summarize(Abundance = sum(Abundance)) %>% 
     ungroup()
```

#### Join melts

```{r}
melts <- list(melt.menu, melt.mp)
```

```{r}
vs <- c('study',
        'subj', 
        'date',
        'detection', 
        'mp_to_menu', 
        'food_group',
        'Abundance') # Shared variables to pull from each phyloseq

lapply(melts, ncol)
melts <- lapply(melts, function(x){select(x, one_of(vs))})
lapply(melts, ncol)
```

```{r}
melt <- bind_rows(melts)
dim(melt)
rm(melts)
```

```{r}
melt$detection <- factor(melt$detection, 
                         levels = c('Menu',
                                    'Protein'))
```

```{r}
# Synchronize food group naming:
melt %>% 
     group_by(detection) %>% 
     summarize(unique(food_group))
```
```{r}
melt$food_group[melt$food_group == 'muscle'] <- 'meat'
melt$food_group[melt$food_group == 'eggs, shell eggs'] <- 'egg'
melt$food_group[melt$food_group == 'poultry'] <- 'meat'
melt$food_group[melt$food_group == 'dairy, solid/semi-solid dairy products'] <- 'dairy'
melt$food_group[melt$food_group == 'dairy, fluid milk'] <- 'dairy'
```

# Analyze

## Menu co-consumption

How many days had recorded co-consumption?

```{r}
melt %>% 
     filter(detection == 'Menu' & Abundance > 0) %>% 
     group_by(study, subj, date, mp_to_menu) %>% 
     summarize(n_tissues = n_distinct(food_group)) %>% 
     ungroup() %>% 
     group_by(mp_to_menu) %>% 
     summarize(co_consumption = sum(n_tissues == 2)/length(n_tissues))
```

Recode as categorical, then do a Chi-squared test:

```{r}
# Does it differ by food type?
x <- 
     melt %>% 
     filter(detection == 'Menu' & Abundance > 0) %>% 
     group_by(study, subj, date, mp_to_menu) %>% 
     summarize(n_tissues = n_distinct(food_group)) %>% 
     mutate(co_consumption = factor(n_tissues,
                                    levels = c(2, 1),
                                    labels = c('Yes', 'No'))) %>% 
     ungroup() %>% 
     select(mp_to_menu, co_consumption) 

x <- table(x$mp_to_menu, x$co_consumption)
x

chisq.test(x, simulate.p.value = TRUE)
```

```{r}
# Does it differ by study?
for (taxon in c('chicken', 'cattle')){
     print(taxon)
     x <- 
          melt %>% 
          filter(mp_to_menu == taxon & detection == 'Menu' & Abundance > 0) %>% 
          group_by(study, subj, date, mp_to_menu) %>% 
          summarize(n_tissues = n_distinct(food_group)) %>% 
          mutate(co_consumption = factor(n_tissues,
                                         levels = c(2, 1),
                                         labels = c('Yes', 'No'))) %>% 
          ungroup() %>% 
          select(study, co_consumption) 
     
     x <- table(x$co_consumption, x$study)
     x
     
     print(chisq.test(x, simulate.p.value = TRUE))
}
```

## Metaproteomic co-consumption

```{r}
# x <- 
     melt |> 
     names()
```


```{r}
x <- melt %>% 
     filter(Abundance > 0) %>% 
     group_by(detection, subj, date, mp_to_menu) %>% 
     summarize(n_tissues = n_distinct(food_group)) %>% 
     mutate(co_consumption = factor(n_tissues,
                                    levels = c(2, 1),
                                    labels = c('Yes', 'No'))) %>% 
     ungroup() 

x <- table(x$mp_to_menu, x$co_consumption)
print(chisq.test(x, simulate.p.value = TRUE))
```

```{r}
for (taxon in c('cattle', 'chicken')){
     print(taxon)
     x <- melt %>% 
          filter(Abundance > 0 & mp_to_menu == taxon) %>% 
          group_by(detection, subj, date, mp_to_menu) %>% 
          summarize(n_tissues = n_distinct(food_group)) %>% 
          mutate(co_consumption = factor(n_tissues,
                                         levels = c(2, 1),
                                         labels = c('Yes', 'No'))) %>% 
          ungroup() 
     
     x <- table(x$detection, x$co_consumption)
     print(x)
     print(chisq.test(x, simulate.p.value = TRUE))
}
```

Oh interesting-- we *are* significantly different for chicken. And we detect significantly more co-consumption. Why could this be?
- Incorrect matching of menu days
- Or, metaproteomic data is a "summary" of several days' worth of intake. 

```{r}
for (taxon in c('cattle', 'chicken')){
     print(taxon)
     x <- melt %>% 
          filter(mp_to_menu == taxon) %>% 
          # Make "non-meat" category for dairy/egg so I can loop over datasets
          mutate(food_group = ifelse(food_group != 'meat',
                                     yes = 'not_meat',
                                     no = 'meat')) %>% 
          group_by(detection, subj, date, mp_to_menu, food_group) %>%
          summarize(Abundance = sum(Abundance)) %>% # Follow up on this-- why dups?
          mutate(Abundance = ifelse(Abundance > 0,
                                    yes = 1,
                                    no = 0)) %>% 
          pivot_wider(names_from = food_group,
                      values_from = Abundance,
                      values_fill = 0) %>% 
          mutate(co_consumption = meat & not_meat,
                 meat_only = meat & !not_meat,
                 not_meat_only = !meat & not_meat) %>% 
          pivot_longer(cols = c(co_consumption,
                                meat_only,
                                not_meat_only),
                       names_to = 'intake_type',
                       values_to = 'value') %>% 
          filter(value == TRUE) 
     
     x <- table(x$detection, x$intake_type)
     print(x)
     print(chisq.test(x), simulate.p.value = TRUE)
}
```

### "Overall" summaries

#### Absolute

Note that this is currently with all menu days included.

```{r}
overall.input <- 
     melt %>% 
     filter(mp_to_menu == 'chicken',
            study == 'Intervention') %>% # Healthy Donor lacking portion size; omit for now
     pivot_wider(names_from = food_group,
                 values_from = Abundance)

overall.input
```

```{r}
ggplot(overall.input, aes(x = meat, y = egg, color = subj)) +
     geom_point(alpha = 0.8, size = 2) +
     facet_wrap(~detection) + 
     coord_fixed() +
     scale_color_manual(values = colors.subj[2:5]) +
     labs(y = 'Egg intake<br>(grams or PSMs)',
          x = 'Meat intake<br>(grams or PSMs)') + 
     theme(legend.position = 'none')
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results', 
            'manuscript',
            'tbd',
            'biomarkers',
            paste0(plotID,
                   '_Chicken meat v. egg intake by detection.pdf')),
       height = 3, width = 8)
```

```{r}
# Note can only do this with pairing
overall.input %>% 
     pivot_longer(cols = c(meat, egg),
                  names_to = 'tissue',
                  values_to = 'Abundance') %>% 
     pivot_wider(names_from = detection,
                 values_from = 'Abundance') %>% 
     ggplot(aes(x = Menu, y = Protein)) +
     geom_point(alpha = 0.8) +
     facet_wrap(~tissue)
```

```{r}
overall.input %>% 
     pivot_longer(cols = c(meat, egg),
                  names_to = 'tissue',
                  values_to = 'Abundance') %>% 
     pivot_wider(names_from = detection,
                 values_from = 'Abundance') %>% 
     group_by(tissue) %>% 
     summarize(rho = cor.test(~ Menu + Protein, method = 'spearman')$estimate,
               p = cor.test(~ Menu + Protein, method = 'spearman')$p.value)
```


#### Proportion

? Will need to ask Manuel if this is valid without adjusting for protein length.

```{r}
overall.input %>% 
     mutate(pct_meat = meat/(meat + egg)) %>% 
     ggplot(aes(x = pct_meat)) +
     # geom_density() +
     geom_histogram(binwidth = 0.1,
                    boundary = 0) +
     facet_wrap(~detection) +
     labs(x = 'Meat %<br>(of meat + egg total)')
```

```{r}
# Can only do this with exact pairing
overall.input %>% 
     mutate(pct_meat = meat/(meat + egg))  %>% 
     select(subj, date, detection, pct_meat) %>% 
     pivot_wider(names_from = detection,
                 values_from = pct_meat) %>% 
     ggplot(aes(x = Menu, y = Protein, color = subj)) +
     geom_point(alpha = 0.8, size = 2) +
     labs(x = 'Menu (% meat by grams)',
          y = 'Protein (% meat by PSMs)') +
     coord_equal()
```

```{r}
overall.input %>% 
     mutate(pct_meat = meat/(meat + egg))  %>% 
     select(subj, date, detection, pct_meat) %>% 
     pivot_wider(names_from = detection,
                 values_from = pct_meat) %>% 
     cor.test(~ Menu + Protein, data = .,
              method = 'spearman')
```


### Pairing to menus

```{r}
taxa <- c('chicken', 'cattle')
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

for (taxon in taxa){
     melt.filt <- filter(melt, common_name == taxon)
     
     ggplot(melt.filt, aes(x = date)) +
          geom_tile(data = samples, aes(x = date, 
                                        y = 0),
                    fill = 'gray90',
                    height = Inf) +
          geom_bar(aes(y = Abundance, fill = detection), stat = 'identity') +
          facet_grid(cols = vars(subj),
                     rows = vars(detection),
                     scales = 'free',
                     space = 'free_x') +
          scale_fill_manual(values = colors.detection) +
          labs(x = 'Date', y = 'Food abundance') +
          theme(axis.text.x = element_blank(),
                legend.position = 'none',
                panel.grid = element_blank())
     
     # ggsave(here('results', 
     #             'biomarkers', 
     #             paste0(plotID, 
     #                    '_',
     #                    taxon,
     #                    ' detection by measure and sample.pdf')),
     #        height = 4)
}
```

### Protein only

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

samples.filt <- filter(samples, detection != 'DNA')

for (taxon in taxa){
     melt.filt <- 
          filter(melt, common_name == taxon & detection != 'DNA')
     
     ggplot(melt.filt, aes(x = date)) +
          geom_tile(data = samples.filt, aes(x = date, 
                                             y = 0),
                    fill = 'gray90',
                    height = Inf) +
          geom_bar(aes(y = Abundance, fill = detection), stat = 'identity') +
          facet_grid(cols = vars(subj),
                     rows = vars(detection),
                     scales = 'free',
                     space = 'free_x') +
          scale_fill_manual(values = colors.detection) +
          labs(x = 'Date', y = 'Food abundance') +
          theme(axis.text.x = element_blank(),
                legend.position = 'none',
                panel.grid = element_blank())
     
     # ggsave(here('results', 
     #             'biomarkers', 
     #             paste0(plotID, 
     #                    '_',
     #                    taxon,
     #                    ' detection by measure and sample.pdf')),
     #        height = 3)
}
```

### Protein by type

#### Cattle 

```{r}
taxon <- 'cattle'
samples.filt <- filter(samples, detection != 'DNA')
melt.filt <- 
     filter(melt, common_name == taxon & detection != 'DNA')
```

##### Recode tissue entries

```{r}
melt.filt %>% 
     group_by(detection) %>% 
     summarize(unique(group))
```

Preserve differences between dairy, but equate meat and muscle.

```{r}
melt.filt$group[melt.filt$group == 'muscle'] <- 'meat'
```

##### Plot

```{r}
colors.cattle <- c(
     '#4e79a7',
     '#4e79a7',
     '#a0cbe8', 
     '#e15759',
     '#bab0ac' # other
                   )
```

Try alternate way of plotting: 1 per subject

```{r}
subjects <- 
     samples %>% 
     pull(subj) %>% 
     unique() %>% 
     sort()

plot.list <- list()

for (s in 1:5){
     if (s != 5){ # Label facet differently in last subject
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.cattle) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     strip.text.y = element_blank(),
                     legend.position = 'none')
     } else {
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         width = 1,
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.cattle) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     legend.position = 'none')
     }
     # Add to plot list
     plot.list[[s]] <- p
}
```

```{r}
# Relative widths:
widths <- 
     melt.filt %>%
     group_by(subj) %>% 
     summarize(dates = n_distinct(date)) %>% 
     ungroup() %>% 
     mutate(relative = dates/sum(dates))

widths
```

```{r}
plot_grid(plotlist = plot.list,
          nrow = 1,
          rel_widths = c(1.2, 2, 2, 2, 2.4))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'biomarkers',
            paste0(plotID,
                   '_',
                   taxon,
                   ' cattle tissue types by sample and menu.pdf')),
       height = 2.5, width = 10)
```

#### Chicken
```{r}
taxon <- 'chicken'
samples.filt <- filter(samples, detection != 'DNA')
melt.filt <- 
     filter(melt, common_name == taxon & detection != 'DNA')
```

##### Recode tissue entries

```{r}
melt %>% 
     group_by(detection) %>% 
     summarize(unique(food_group))
```

```{r}
melt.filt$group[melt.filt$group == 'muscle'] <- 'meat'
melt.filt$group[melt.filt$group == 'poultry'] <- 'meat'
melt.filt$group[melt.filt$group == 'eggs, shell eggs'] <- 'egg'
```

```{r}
colors.chicken <- c('#f1ce63',
                   '#ff9d9a',
                   '#bab0ac')
```

```{r}
subjects <- 
     samples %>% 
     pull(subj) %>% 
     unique() %>% 
     sort()

plot.list <- list()

for (s in 1:5){
     if (s != 5){ # Label facet differently in last subject
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.chicken) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     strip.text.y = element_blank(),
                     legend.position = 'none')
     } else {
          p <- melt.filt %>% 
               filter(subj == subjects[[s]]) %>% 
               ggplot(aes(x = date, 
                          y = Abundance, 
                          fill = group)) +
               geom_tile(data = filter(samples.filt, 
                                       subj == subjects[[s]]), 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         width = 1,
                         height = Inf) +
               geom_bar(stat = 'identity') +
               labs(fill = 'Tissue') +
               facet_grid(cols = vars(subj), 
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.chicken) +
               theme(axis.text.x = element_blank(),
                     axis.ticks.x = element_blank(),
                     axis.title.x = element_blank(),
                     axis.title.y = element_blank(),
                     panel.grid = element_blank(),
                     legend.position = 'none')
     }
     # Add to plot list
     plot.list[[s]] <- p
}
```

```{r}
# Relative widths:
widths <- 
     melt.filt %>%
     group_by(subj) %>% 
     summarize(dates = n_distinct(date)) %>% 
     ungroup() %>% 
     mutate(relative = dates/sum(dates))

widths
```

```{r}
plot_grid(plotlist = plot.list,
          nrow = 1,
          rel_widths = c(1.2, 2, 2, 2, 2.4))
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'biomarkers',
            paste0(plotID,
                   '_',
                   taxon,
                   ' chicken tissue types by sample and menu.pdf')),
       height = 2.5, width = 10)
```