---
title: "Biomarkers"
output: html_notebook
---

Only want to use menu and metaproteomic data, and consider the best-synchronized name between the two

## Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, echo=FALSE}
library(cowplot)
library(here) 
library(phyloseq) 
library(tidyverse) 
```

```{r}
colors.detection <- 
     c('#fbb04e',
       '#7e756d')

names(colors.detection) <- 
     c('Protein',
       'Menu')
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title.x = ggtext::element_markdown(size = 14),
                    axis.title.y = ggtext::element_markdown(size = 14),
                    legend.title = element_text(size = 12),
                    strip.text = element_text(size = 12)
               )
)
```

## Read in data

### Taxon name mapping

```{r}
name.sync <- 
     here('data',
          'processed',
          'Naming discrepancies.xlsx') %>% 
     readxl::read_excel()

name.sync
```

### Sample dates

```{r}
samples <- 
     here('data',
          'metadata',
          'Sample collection times.csv') %>% 
     read_csv()

samples
```

### Metaproteomic phyloseq

```{r}
ps.protein <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          # 'ps_mp_1PUP.rds') %>% # >=1 protein unique peptide
          'ps_mp_1UP.rds') %>% # >=1 unique peptide
     readRDS()
          
ps.protein
```
### Protein annotations

```{r}
# Raw protein list
proteins.df <- 
     here('data',
          'processed',
          '5FDR_1UP_both datasets.rds') %>% 
     readRDS()
```

```{r}
annotations <- 
     here('data',
          'processed',
          'Animal-based proteins (full annotation).csv') %>% 
     read_csv()

annotations
```
## Pre-process

### Samples

```{r}
# Tidy variables
samples$date <- as.Date(samples$date,
                        format = '%m/%d/%y')

samples$subj <- factor(samples$subj, 
                       labels = c('HD',
                                  'ID1', 'ID2', 'ID3', 'ID4'))

samples
```

### Names

```{r}
# Remove taxa where distinction is made in menu to prevent duplicate join
name.sync <- 
     filter(name.sync, 
            (menu_taxa != 'Pisum sativum var. macrocarpum' | is.na(menu_taxa)))
```

### Proteins

Filter out proteins from animal taxa without an annotation ("other"), which may include host- and microbial cross-identifications.

```{r}
exclude <- 
     annotations |> 
     filter(category == 'other') |> 
     pull(word)

# Now find corresponding protein accessions
nrow(proteins.df)
proteins.df <-
     proteins.df |> 
     mutate(word = gsub(Description,
                        pattern = ' OS=.*',
                        replacement = '')) |> 
     filter(!(word %in% exclude))

nrow(proteins.df)
```

Note that the dilemma here is I'm curating this for animals only-- no similar filter is being applied to plants and fungi.

```{r}
# Also remove these entries from phyloseq object before melting
ntaxa(ps.protein)

ps.protein <- 
     prune_taxa(taxa_names(ps.protein) %in% proteins.df$Accession,
                ps.protein)

ntaxa(ps.protein)
```

### Melted phyloseq

```{r}
melt <- psmelt(ps.protein)
```

```{r}
# Join to common names
melt <- 
     name.sync |> 
     filter(!is.na(mp_taxa)) |> 
     select(mp_taxa, common_names) |> 
     right_join(melt, 
                by = join_by(mp_taxa == name),
                multiple = 'all') |> 
     select(OTU,
            kingdom,
            name = mp_taxa,
            common_name = common_names, # Not all filled in
            # May need to revisit
            Sample,
            Abundance,
            study,
            subj,
            date)
```

### >=5 detections

```{r}
multiples <- 
     melt |> 
     group_by(kingdom, name, common_name, Sample) |> 
     summarize(PSMs = sum(Abundance)) |> 
     filter(PSMs > 0) |> 
     group_by(kingdom, name, common_name) |> 
     summarize(detections = n_distinct(Sample)) |> 
     filter(detections >= 5)

arrange(multiples, desc(detections))
```

```{r}
# Filter data to only these taxa
dim(melt)
melt <- 
     filter(melt,
            name %in% multiples$name)

dim(melt)
```

Interesting-- this removes less than I expected.

## Analyze

```{r}
# What's our starting potential number of proteins?
n_distinct(melt$OTU)
```


Try to identify "dominant proteins" for each taxon. 

One simple theory is that there's more likely to be a dominant protein if there are fewer proteins overall.

So I could tally those and use them as a starting point.

```{r}
multiples <- 
     melt |> 
     group_by(name) |> 
     summarize(unique_PSMs = n_distinct(OTU),
               cumulative_PSMs = sum(Abundance),
               ratio = cumulative_PSMs/unique_PSMs) |> 
     left_join(x = multiples) |> 
     arrange(unique_PSMs, detections)

multiples
```

```{r}
# Write to table
# write_csv(multiples,
#           here('data',
#                'processed',
#                'biomarkers',
#                'Automated biomarker list (taxon-level, >=5 detections).csv'))
```

## Visualize

### Cumulative v unique PSMs

#### Scatterplot

```{r}
ggplot(multiples, aes(x = unique_PSMs,
                      y = cumulative_PSMs,
                      size = detections,
                      color = kingdom)) +
     geom_point(alpha = 0.3) +
     scale_color_manual(values = c('gray50', '#e15759', '#59a14f')) + 
     scale_y_log10() +
     labs(y = 'Total PSMs identified to food',
          x = 'Number of unique proteins<br>detected for food',
          size = 'Number of samples\nfood present in',
          color = 'Food kingdom')
     
```

```{r}
ggplot(multiples, aes(x = unique_PSMs,
                      y = cumulative_PSMs,
                      size = detections,
                      color = kingdom)) +
     geom_point(alpha = 0.3) +
     scale_color_manual(values = c('gray50', '#e15759', '#59a14f')) + 
     scale_y_log10() + 
     scale_x_log10() +
     labs(y = 'Total PSMs identified to food',
          x = 'Number of unique proteins<br>detected for food',
          size = 'Number of samples\nfood present in',
          color = 'Food kingdom')
```

This looks like a pretty reasonable threshold-- we'll keep everything less than ~32 unique PSMs as a manageable first pass (10^1.5).

```{r}
# How many proteins would be kept by this strategy?
# Make a list of brief annotations for joining
annotations_brief <- 
     proteins.df |> 
     select(Accession, Description) |> 
     distinct()

# And a reduced melt for joining too
melt_brief <- 
     melt |> 
     select(OTU, 
            name,
            Abundance) |>
     group_by(OTU, name) |> 
     summarize(Abundance = sum(Abundance))

uniques <- 
     multiples |> 
     filter(unique_PSMs < 32) |> 
     left_join(melt_brief,
               multiple = 'all') |> 
     left_join(annotations_brief,
               by = c('OTU' = 'Accession')) |> 
     select(-ends_with('PSMs'),
            -ratio)
```

#### Ratio histogram

Another way to get at cutpoint
A neat supplemental plot might be labeling these outliers
Or perhaps labeling them on the scatterplot above

```{r}
ggplot(multiples, aes(x = ratio)) +
     geom_histogram(boundary = 0,
                    binwidth = 5)
```

## Shared across samples

Now try filtering to only those that are shared across every sample with a detection
The concern here is that if false-positive detections (homology to host or microbe)-- would also expect to turn up here too.

```{r}
shared <- 
     select(multiples,
            kingdom, 
            name,
            common_name,
            taxon_samples = detections) # Total samples taxon was detected in
     
shared <- 
     melt |> 
     filter(Abundance > 0) |> 
     group_by(name, OTU) |> 
     summarize(protein_samples = n_distinct(Sample), # Total protein was detected in
               total_PSMs = sum(Abundance)) |>  
     ungroup() |> 
     right_join(shared, 
                multiple = 'all')
```

Now just as a first pass let's see if any of these are identical (i.e. the taxon was detected in 10 samples and the specific protein was also detected in 10 samples)

```{r}
shared |> 
     filter(protein_samples == taxon_samples) |> 
     left_join(annotations_brief,
               by = c('OTU' = 'Accession')) |> 
     select(name, common_name, OTU, Description)
```
```{r}
# >50% of detections
shared <-
     shared |> 
     filter(protein_samples >= taxon_samples*0.5) |> 
     left_join(annotations_brief,
               by = c('OTU' = 'Accession')) |> 
     select(name, 
            common_name, 
            OTU, 
            Description, 
            n_samples_with_taxon = taxon_samples,
            n_samples_with_protein = protein_samples,
            total_PSMs)
```

```{r}
write_csv(shared,
          here('data', 
               'processed',
               'biomarkers',
               'Automated biomarker list (shared, >=5 detections).csv'))
```


### Scatterplot

```{r}
ggplot(shared, aes(x = taxon_samples, y = protein_samples)) + 
     geom_abline(slope = 1,
                 intercept = 0,
                 color = '#e15759',
                 size = 1,
                 linetype = 'dashed') +
     geom_point(alpha = 0.1, size = 3) +
     xlim(1, 26) +
     ylim(1, 26) +
     coord_equal()
```

What gets included at cutoff by percentage?

```{r}
ggplot(shared, aes(x = taxon_samples, 
                   y = protein_samples,
                   color = (protein_samples/taxon_samples) > 0.5)) + 
     geom_abline(slope = 1,
                 intercept = 0,
                 color = '#e15759',
                 alpha = 0.8,
                 size = 1,
                 linetype = 'dashed') +
     geom_point(alpha = 0.3, size = 3) +
     xlim(1, 26) +
     ylim(1, 26) +
     coord_equal() +
     scale_color_manual(values = c('gray80', '#59a14f')) +
     labs(x = 'Samples with detected food',
          y = 'Samples with detected protein<br>(from same food)',
          color = 'Protein detected in\n>50% of cases')
```

```{r}
# Filter and save this list

```

## Intersection of the two approaches



### Apple

Try for one food first

```{r}
taxon <- 'carrot'
```

Stream plot of intake

```{r}
melt.filt <- filter(melt, mp_to_menu == taxon)
     
ggplot(melt.filt, aes(x = date)) +
     geom_tile(data = samples, aes(x = date, 
                                   y = 0),
               fill = 'gray90',
               height = Inf) +
     geom_bar(aes(y = Abundance, fill = detection), stat = 'identity') +
     facet_grid(cols = vars(subj),
                rows = vars(detection),
                scales = 'free',
                space = 'free_x') +
     scale_fill_manual(values = colors.detection) +
     labs(x = 'Date', y = 'Food abundance') +
     theme(axis.text.x = element_blank(),
           legend.position = 'none',
           panel.grid = element_blank())
```
What proteins do we detect (stacked bar of contributions)? 

```{r}
proteins.df %>% 
     filter(common_name == 'apple') %>% 
     ggplot(aes(x = id)) +
     geom_bar(aes(y = `# PSMs`, fill = Accession), 
              stat = 'identity',
              position = 'stack') +
     facet_wrap(~subj,
                scales = 'free_x') +
     labs(x = 'Date', y = 'PSMs') +
     theme(axis.text.x = element_blank(),
           legend.position = 'none',
           panel.grid = element_blank())
```
What proteins are most consistently detected?

```{r}
# How many unique accessions for apple? 
proteins.df %>% 
     filter(common_name == 'apple') %>% 
     pull(Accession) %>% 
     n_distinct()
```

```{r}
proteins.df %>% 
     filter(common_name == 'apple') %>% 
     group_by(Accession) %>% 
     summarize(total_psms = sum(`# PSMs`),
               prevalence = n_distinct(id)) %>% 
     arrange(desc(total_psms))
```
```{r}
# What are these top four?
top4 <- 
     proteins.df %>% 
     filter(common_name == taxon) %>% 
     group_by(Accession) %>% 
     summarize(total_psms = sum(`# PSMs`),
               prevalence = n_distinct(id)) %>% 
     slice_max(total_psms, n = 4) %>% 
     left_join(select(proteins.df, 
                      Accession,
                      Description)) %>% 
     pull(Description) %>% 
     unique()

top4
```

```{r}
# Try highlighting on plots
melt.filt <- filter(melt, mp_to_menu == taxon)
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

for (i in 1:4){
     proteins.df %>% 
     filter(common_name == taxon) %>% 
     ggplot(aes(x = id)) +
     geom_bar(aes(y = `# PSMs`, fill = Description == top4[i]), 
              stat = 'identity',
              position = 'stack') +
     facet_wrap(~subj,
                scales = 'free_x',
                nrow = 1) +
     labs(x = 'Date', y = 'PSMs') +
     theme(axis.text.x = element_blank(),
           legend.position = 'none',
           panel.grid = element_blank())
     
     ggsave(here('results', 
                 'manuscript',
                 'tbd',
                 'biomarkers',
            paste0(plotID, '_', taxon, 'top ', i, ' prevalent protein.pdf')),
       height = 4, width = 6)
}
```

