---
title: "Estimating individualized transit times"
author: "Brianna Petrone"
output: html_notebook
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(cowplot)
library(eulerr) # For Euler plots
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
colors.detection <- 
     c('#849db1', 
       '#fbb04e',
       '#7e756d')

names(colors.detection) <- 
     c('DNA',
       'Protein',
       'Menu')
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title.x = ggtext::element_markdown(size = 14),
                    axis.title.y = ggtext::element_markdown(size = 14),
                    legend.title = element_text(size = 12,
                                                face = 'bold'),
                    strip.text = element_text(size = 12,
                                              face = 'bold')
               )
)
```

## Read in files

### Sample dates

```{r}
samples <- 
     here('data',
          'metadata',
          'Sample collection times.csv') %>% 
     read_csv()

samples
```

### Taxon name mapping

```{r}
name.sync <- 
     here('data',
          'processed',
          'Naming discrepancies.xlsx') %>% 
     readxl::read_excel()

name.sync
```

### Metaproteomic

```{r}
ps.protein <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          # 'ps_mp_1PUP.rds') %>% # >=1 protein unique peptide
          'ps_mp_1UP.rds') %>% # >=1 unique peptide
     readRDS()
          
ps.protein
```

### Menu

```{r}
ps.menu <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_menu.rds') %>% 
     readRDS()

ps.menu
```

### Metabarcoding phyloseqs


```{r}
ps.mb.plant <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_trnL.rds') %>% 
     readRDS()

ps.mb.plant
```

```{r}
ps.mb.animal <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_12SV5.rds') %>% 
     readRDS()

ps.mb.plant
```

## Pre-process

### Samples

```{r}
samples$date <- as.Date(samples$date,
                        format = '%m/%d/%y')
```

```{r}
# Add detection for each so shading is done by facet, below
detection <- c('Menu', 'DNA', 'Protein')
subj <- unique(samples$subj)

samples <-
     expand.grid(detection = detection, 
                 subj = subj) %>% 
     right_join(samples) %>% 
     arrange(study, subj, date) %>% 
     select(subj:time, detection)

samples
```

### Melt 
```{r}
melt.menu <- psmelt(ps.menu) 
melt.mb.plant <- psmelt(ps.mb.plant)
melt.mb.animal <- psmelt(ps.mb.animal)
melt.mp <- psmelt(ps.protein)
```

#### Add common name
Add common name and group by dataset:

```{r}
melt.menu <- 
     melt.menu %>% 
     left_join(select(name.sync,
                      menu_taxa, common_name),
               by = c('OTU' = 'menu_taxa'))

melt.mb.animal <- 
     melt.mb.animal %>% 
     left_join(select(name.sync,
                      mb_taxa, common_name),
               by = c('OTU' = 'mb_taxa'))

melt.mb.plant <- 
     melt.mb.plant %>% 
     left_join(select(name.sync,
                      mb_taxa, common_name),
               by = c('OTU' = 'mb_taxa'))

# Note column naming differences for metaproteomics
melt.mp <- 
     melt.mp %>% 
     left_join(select(name.sync,
                      mp_taxa, common_name),
               by = c('name' = 'mp_taxa'))
```

#### Add detection method

```{r}
# Add dataset info to each melted object
# melt.menu$detection <-  'Menu'
melt.menu$detection <- 'Menu'
melt.mb.plant$detection <- 'DNA'
melt.mb.animal$detection <- 'DNA'
melt.mp$detection <- 'Protein'
```

#### Join melts

```{r}
melts <- list(
     melt.menu, # Full menu
     melt.mb.animal,
     melt.mb.plant,
     melt.mp)
```

```{r}
vs <- c('study',
        'subj', 
        'date',
        'detection', 
        'common_name', 
        'Abundance') # Shared variables to pull from each phyloseq

lapply(melts, ncol)
melts <- lapply(melts, function(x){select(x, one_of(vs))})
lapply(melts, ncol)
```

```{r}
melt <- bind_rows(melts)
dim(melt)
rm(melts)
```

```{r}
melt$detection <- factor(melt$detection, 
                         levels = c('Menu',
                                    'DNA',
                                    'Protein'))
```

## Analyze

### Find uniquely detected foods

#### Count

```{r}
uniques <- 
     melt %>% 
     group_by(detection, subj, common_name) %>% 
     filter(Abundance > 0) %>% 
     count() %>% 
     filter(n == 1) %>% 
     ungroup() %>% 
     select(-n)

uniques
```
```{r}
# Plot
uniques %>% 
     group_by(subj, detection) %>% 
     count() %>% 
     ggplot(aes(x = subj, y = n, fill = detection)) +
     geom_bar(stat = 'identity',
              position = 'dodge') +
     labs(y = 'Foods detected or<br>consumed only once',
          x = 'Participant') +
     scale_fill_manual(values = colors.detection) +
     theme(legend.title = element_blank(),
           panel.grid = element_blank())
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 
#             'transit time',
#             paste0(plotID, '_Unique foods by participant and method.pdf')),
#        height = 3, width = 5)
```

```{r}
# Median and MAD to add to plots
uniques.stat <- 
     uniques %>% 
     group_by(subj, detection) %>% 
     count() %>%  
     group_by(detection) %>% 
     summarize(med = median(n),
               mad = mad(n))

uniques.stat
```

```{r}
# Test for difference
# Think this is most appropriately a Friedman test: "Similar to the parametric repeated measures ANOVA, it is used to detect differences in treatments across multiple test attempts"

uniques.test <- 
     uniques %>% 
     group_by(detection, subj) %>% 
     count() %>% 
     ungroup() %>% 
     pivot_wider(names_from = 'detection',
                 values_from = 'n') %>% 
     column_to_rownames(var = 'subj') %>% 
     as.matrix()

dimnames(uniques.test)

friedman.test(uniques.test)
PMCMRplus::frdAllPairsNemenyiTest(uniques.test)
```

```{r}
# Plot
uniques %>% 
     group_by(subj, detection) %>% 
     count() %>% 
     ggplot(aes(x = detection)) +
     geom_jitter(aes(y = n, color = detection),
                 alpha = 0.8,
                 height = 0,
                 width = 0.05) +
     geom_errorbar(data = uniques.stat,
                   aes(ymin = med-mad,
                       ymax = med+mad),
                   width = 0.05) +
     geom_errorbar(data = uniques.stat,
                    aes(ymin = med,
                        ymax = med),
                    width = 0.2) +
     labs(y = 'Foods detected or<br>consumed only once',
          x = 'Method') +
     scale_color_manual(values = colors.detection) +
     ylim(0, 28) +
     theme(legend.position = 'none',
           axis.text.x = element_blank(),
           axis.ticks.x = element_blank())
```
```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 
#             'transit time',
#             paste0(plotID, '_Unique foods by method.pdf')),
#        height = 3,
#        width = 2)
```

Follow-on thoughts: 
* How believable is the signal in each case? 
     + Is it abundant? (High number of PSMs)?
     + Is it backed up by menu data/trnL? (Menu data would be the necessary feature to get at transit time).
     
#### Intersection

How many are in common across the three datasets?

```{r}
# Reformat for Euler plots
uniques.wide <-
     uniques %>% 
     mutate(detected = 1) %>% 
     pivot_wider(names_from = 'detection',
                 values_from = 'detected',
                 values_fill = 0) 

uniques.wide
```

```{r}
subj1 <- 
     uniques.wide %>% 
     filter(subj == '908') %>% 
     select(DNA, Protein, Menu) %>% 
     euler() %>% 
     plot(
          fills = colors.detection,
          labels = FALSE,
          shape = "ellipse", 
          quantities = TRUE
          )

subj2 <- 
     uniques.wide %>% 
     filter(subj == 'AK65') %>% 
     select(DNA, Protein, Menu) %>% 
     euler() %>% 
     plot(
          fills = colors.detection,
          labels = FALSE,
          shape = "ellipse", 
          quantities = TRUE
          )

subj3 <- 
     uniques.wide %>% 
     filter(subj == 'JN55') %>% 
     select(DNA, Protein, Menu) %>% 
     euler() %>% 
     plot(
          fills = colors.detection,
          labels = FALSE,
          shape = "ellipse", 
          quantities = TRUE
          )

subj4 <- 
     uniques.wide %>% 
     filter(subj == 'ML48') %>% 
     select(DNA, Protein, Menu) %>% 
     euler() %>% 
     plot(
          fills = colors.detection,
          labels = FALSE,
          shape = "ellipse", 
          quantities = TRUE
          )

subj5 <- 
     uniques.wide %>% 
     filter(subj == 'PC74') %>% 
     select(DNA, Protein, Menu) %>% 
     euler() %>% 
     plot(
          fills = colors.detection,
          labels = FALSE,
          shape = "ellipse", 
          quantities = TRUE
          )
```

```{r}
plot_grid(subj1, subj2, subj3, subj4, subj5,
          nrow = 5,
          scale = 0.9) + # To give a bit more margin between plots
     draw_plot_label(label = c('  908', 
                               'AK65', 
                               'JN55',
                               'ML48',
                               'PC74'),
                     x = -0.05,
                     y = .21*4:0 + 0.09,
                     size = 12) +
     theme(plot.margin = unit(c(0.5, 0.5, 0.5, 0.5), "cm")) 
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

# ggsave(here('results', 
#             'transit time',
#             paste0(plotID, '_Intersection of uniquely identified foods.pdf')),
#        height = 6,
#        width = 4)
```

#### Abundance

```{r}
# Join back to abundance data from big melt object
uniques <- 
     melt %>% 
     select(names(uniques), Abundance) %>% 
     filter(Abundance > 0) %>% 
     left_join(uniques, .)
```

Make one plot per subject, and aggregate.  Prioritize by menu items that are most abundant:
```{r}
subjects <- sort(unique(melt$subj))

plot.list <- list()

for (s in 1:5){
     uniques.subj <- 
          uniques %>% 
          filter(subj == subjects[s] & detection == 'Menu') %>% 
          arrange(desc(Abundance), common_name) %>% 
          mutate(common_name = factor(common_name,
                                      levels = common_name))
     
     uniques.subj$common_name <- factor(uniques.subj$common_name,
                                        levels = uniques.subj$common_name)
          
     p <- ggplot(uniques.subj,
                 aes(x = common_name, 
                     y = Abundance, 
                     fill = detection)) +
          geom_bar(stat = 'identity') +
          scale_fill_manual(values = colors.detection['Menu']) +
          theme(axis.title.x = element_blank(),
                axis.title.y = element_blank(),
                axis.text.x = element_text(size = 6,
                                           angle = 30, 
                                           hjust = 1),
                legend.position = 'none',
                panel.grid = element_blank())
     
     if (subjects[s] != '908'){
          # Add ~1 oz amount threshold 
          p <- p + geom_hline(yintercept = 28, color = '#e03531')
     }
     
     plot.list[[s]] <- p
}
```

```{r}
plot_grid(plotlist = plot.list,
          nrow = 5) +
     theme(plot.margin = unit(c(0, 0, 0, 0.5), 'cm'))
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
# ggsave(here('results',
#             'transit time',
#             paste0(plotID,
#                    '_Unique menu items by abundance and subject.pdf')),
#        height = 10, width = 4)
```

#### Time series

```{r}
# Here, might as well try uniquely detected foods by any source, not just menu. Make infrastructure for iterating over subjects and foods:
foods.908 <- 
     uniques %>% 
     filter(subj == '908') %>% 
     pull(common_name) %>% 
     unique()

foods.AK65 <- 
     uniques %>% 
     filter(subj == 'AK65') %>% 
     pull(common_name) %>% 
     unique()

foods.JN55 <- 
     uniques %>% 
     filter(subj == 'JN55') %>% 
     pull(common_name) %>% 
     unique()

foods.ML48 <- 
     uniques %>% 
     filter(subj == 'ML48') %>% 
     pull(common_name) %>% 
     unique()

foods.PC74 <- 
     uniques %>% 
     filter(subj == 'PC74') %>% 
     pull(common_name) %>% 
     unique()
```

```{r}
foods <- list(foods.908,
              foods.AK65,
              foods.JN55,
              foods.ML48,
              foods.PC74)

names(foods) <- subjects
```

```{bash}
cd results/transit\ time
mkdir 908
mkdir AK65
mkdir JN55
mkdir ML48
mkdir PC74
```

```{r}
# Iterate over all, saving plots
for (i in seq_along(foods)){
     subject <- names(foods)[i]
     subject.foods <- foods[[i]]
     
     samples.filt <- filter(samples, subj == subject)
     
     # Set plotID by subject
     plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
     
     for (food in subject.foods){
          melt.filt <- 
               filter(melt, subj == subject & common_name == food)
          
          ggplot(melt.filt, aes(x = date)) +
               geom_tile(data = samples.filt, 
                         aes(x = date, y = 0),
                         fill = 'gray90',
                         width = 1,
                         height = Inf) +
               geom_bar(aes(y = Abundance, fill = detection), 
                        stat = 'identity',
                        width = 1) + 
               facet_grid(cols = vars(common_name),
                          rows = vars(detection),
                          scales = 'free',
                          space = 'free_x') +
               scale_fill_manual(values = colors.detection) +
               scale_x_date(date_breaks = '1 day') +
               labs(x = 'Date', y = 'Food abundance') +
               theme(axis.text.x = element_blank(),
                     legend.position = 'none',
                     panel.grid = element_blank())
          
          # Save plot
          ggsave(here('results',
                      'transit time',
                      subject,
                      paste0(plotID,
                             paste0('_', subject,'_', food),
                             '_time series.pdf')),
                 height = 3, width = 4)
     }
}
```


### Unique in metaproteomics, menu

```{r}
intersect.menu.protein <- 
     full_join(uniques.menu, uniques.protein,
          by = c('subj', 'food')) %>% 
     select(subj, food,
            menu_abd = Abundance.x, psm_abd = Abundance.y,
            menu_sample = Sample.x, psm_sample = Sample.y) %>% 
     arrange(subj, desc(psm_abd))
     
intersect.menu.protein %>% 
     filter(!is.na(menu_abd) & !is.na(psm_abd))
```
So overall, this is encouraging in that the detected PSMs almost always come after the food was recorded as consumed in the menu.  However, this seems only to work for PSMs that are detected in very low absolute amounts.  Can I do a manual comparison to see if names aren't matching up?

#### Manual notes

##### AK65

*Sesamum indicum*: detected in stool on 2019/09/26.  In menu, can occur from sesame oil, sesame seeds, or tahini.

```{r}
ps.menu %>% 
     subset_taxa(genus == 'Sesamum') %>% 
     subset_samples(subj == 'AK65') %>% 
     prune_samples(sample_sums(.) > 0, .) %>% 
     otu_table()
```

9/17: Garlicky Sesame Broccoli (sesame oil)
9/18: Spicy Stir-Fried Snow Peas (toasted sesame seeds), totaling 1 g dry weight
9/24: Sesame Snap Peas w/Carrots and Peppers (sesame oil, toasted sesame seeds); Sesame-Shiitake Bok Choy (sesame oil, toasted sesame seeds). Seeds totaling 1.467 g dry weight
9/27: Vietnamese Chicken Meatballs (sesame oil)

*Apium graveolens*: detected in stool on 2019/09/27

```{r}
ps.menu %>% 
     subset_taxa(species == 'Apium graveolens') %>% 
     subset_samples(subj == 'AK65') %>% 
     prune_samples(sample_sums(.) > 0, .) %>% 
     otu_table()
```

9/26: Radish, Celery & Snap Pea Salad (raw celery)

*Oreochromis niloticus* (tilapia): Interestingly, recorded only once in menu (9/23), but turns up with bimodal distribution in stool samples:

```{r}
ps.protein %>% 
     subset_samples(subj == 'AK65') %>% 
     subset_taxa(genus == 'Oreochromis') %>% 
     otu_table() %>% 
     data.frame(PSMs = .) %>% 
     rownames_to_column(var = 'sample') %>% 
     mutate(date = as.Date(gsub('AK65_', '', sample))) %>% 
     ggplot(aes(x = date, y = Oreochromis.niloticus)) +
     geom_bar(stat = 'identity') +
     geom_vline(xintercept = as.Date('2019-09-23'),
                color = 'red', alpha = 0.5, linetype = 'dashed') +
     labs(y = 'PSMs') +
     theme_bw() +
     theme(axis.title.x = element_blank())
```


### Unique in metabarcoding, menu


```{r}

```


### Unique in metabarcoding, metaproteomics