---
title: "Figure 3"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, echo=FALSE}
library(here) 
library(phyloseq) 
library(tidyverse) 
```

# Read in data

## Synchronized names

```{r}
name.sync <- 
     here('data',
          'processed',
          'Naming discrepancies.xlsx') %>% 
     readxl::read_excel()

name.sync
```

## Menu data phyloseq

```{r}
# Menu data phyloseq
ps.menu <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_menu.rds') %>% 
     readRDS()

ps.menu
```
## Metabarcoding phyloseqs

```{r}
# Metabarcoding phyloseqs
ps.animal <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_12SV5.rds') %>% 
     readRDS()

ps.animal

ps.plant <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_trnL.rds') %>% 
     readRDS()

ps.plant
```
## Metaproteomic phyloseq

```{r}
ps.protein <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          # 'ps_mp_1PUP_taxa.rds') %>% # >=1 protein unique peptide
          'ps_mp_1UP_taxa.rds') %>% # >=1 unique peptide
     readRDS()
          
ps.protein
```

# Pre-process

## Menu

```{r}
# This currently has food group info in taxon name:
ps.menu |> 
     taxa_names() |> 
     head()
```
```{r}
# Group by taxon to omit
ps.menu <- 
     ps.menu %>% 
     tax_glom(taxrank = 'name')

# Update names, which previously included food group
taxa_names(ps.menu) <- tax_table(ps.menu)@.Data[, 'name']

ps.menu |> 
     taxa_names() |> 
     head()
```

## Metabarcoding

```{r}
# The 12SV5 data have a shifted taxonomy, maybe due to assignTaxonomy()?
# Below, kingdom should be superkingdom, phylum should be kingdom.
# Think it's class that's throwing this off, for some should be Mammalia
ps.animal@tax_table@.Data
```
```{r}
# Update accordingly
tax_table(ps.animal) <- 
     tax_table(ps.animal)@.Data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     mutate(superkingdom = kingdom,
            kingdom = phylum,
            phylum = class) %>% 
     select(-class) %>% 
     select(superkingdom, kingdom, phylum, everything()) %>% 
     column_to_rownames(var = 'row') %>% 
     as.matrix()

tax_table(ps.animal)[1:3, 1:3]
```

```{r}
# Note failed trnL samples and relabel with NAs prior to merge
failed <- 
     which(sample_sums(ps.plant)==0) %>% 
     names()

asvtab.plant <- ps.plant@otu_table@.Data
asvtab.plant[failed, ] <- NA
otu_table(ps.plant) <- otu_table(asvtab.plant,
                                 taxa_are_rows = FALSE)

# Combine into a single object
ps.mb <- merge_phyloseq(ps.plant, ps.animal)
ps.mb

# Remove separated objects
rm(ps.plant, ps.animal)
```
## Metaproteomics

Nothing to do here:

```{r}
taxa_names(ps.protein) |> head()
ps.protein
```
## Pair with 1-2 days prior intake

For this broadest level comparison, let's look only at the two days prior to sampling.

```{r}
# These already exist in the sample data:
sample_variables(ps.mb)[grep(pattern = 'delta',
                             sample_variables(ps.mb))]

sample_variables(ps.protein)[grep(pattern = 'delta',
                             sample_variables(ps.protein))]
```

```{r}
samdf.mb <- 
     sample_data(ps.mb) %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') 
```

```{r}
# Collapse the ASV table entries from the two days prior to each sample into a single entry.  NOTE: Can't use merge_samples on phyloseq here because some samples make up >1 merge

# Do manually on ASV table
asvtab.menu <- otu_table(ps.menu)@.Data
asvtab.menu.12 <- data.frame()

for (sample in seq(dim(samdf.mb)[1])){
     # Pull menu day -1
     one.before <- asvtab.menu[samdf.mb$delta1[sample], ]
     # Pull menu day -2
     two.before <- asvtab.menu[samdf.mb$delta2[sample], ]
     # Merge and place in new, aggregated OTU table
     asvtab.menu.12 <- rbind(asvtab.menu.12,
                             one.before + two.before)
     # Update food names (only has to be done once)
     if (sample == 1){names(asvtab.menu.12) <- names(one.before + two.before)}
     # Update sample name
     row.names(asvtab.menu.12)[sample] <- samdf.mb$row[sample]
}

# Now rebuild a subsetted phyloseq object
ps.menu.12 <- phyloseq(otu_table(asvtab.menu.12, taxa_are_rows = FALSE),
                       sample_data(ps.mb), # Now this matches
                       tax_table(ps.menu)) # Menu-specific taxonomy

# Remove any taxa that aren't present any longer
ps.menu.12 <- prune_taxa(taxa_sums(ps.menu.12) > 0, ps.menu.12)
ps.menu.12
```
```{r}
# How many taxa are present in each of the objects about to be compared?
cat('Recorded:', ntaxa(ps.menu.12), '\n')
cat('DNA:', ntaxa(ps.mb), '\n')
cat('Protein:', ntaxa(ps.protein), '\n')
```

## Reference taxa by their shared index

Will need to glom by the shared taxon name, because otherwise phyloseq object would have multiple instances of that taxon. 

So what I'll want to do here is join the taxon indices to the names to the taxonomy, and then glom there and rename in the object.

Here, use "mb_to_mp_to_menu" column, which maps between all three datasets.

### Update diet record names

```{r}
# Pull taxtab, dropping food group column
tax_table(ps.menu.12) <- 
     ps.menu.12@tax_table@.Data[, c("superkingdom",
                                    "kingdom",
                                    "phylum",
                                    "order",
                                    "family",
                                    "genus",
                                    "species",
                                    "subspecies",
                                    "varietas",
                                    "name")] 

# Update name
tax_table(ps.menu.12) <- 
     name.sync %>% 
     select(mb_to_mp_to_menu, menu_taxa) %>% 
     right_join(data.frame(ps.menu.12@tax_table@.Data), 
               by = c('menu_taxa' = 'name')) %>% 
     distinct() %>% 
     column_to_rownames(var = 'menu_taxa') %>% 
     as.matrix()
```

```{r}
# Get a list of duplicates that result from metabarcoding data not identifying all taxa to the species level
dups <- 
     ps.menu.12@tax_table@.Data %>% 
     data.frame() %>% 
     filter(duplicated(mb_to_mp_to_menu)) %>% 
     pull(mb_to_mp_to_menu) %>% 
     unique()

dups
```
```{r}
# Merge these manually
ps.menu.12
merges = list()
for (taxon in dups){
     # Get the list of equivalent taxa at the species level for that food
     eq_taxa <- 
          ps.menu.12@tax_table@.Data %>% 
          data.frame() %>% 
          filter(mb_to_mp_to_menu == taxon) %>% 
          select(-mb_to_mp_to_menu) %>% 
          MButils::lowest_level() %>% 
          pull(name)
     
     merges <- append(merges, list(eq_taxa))
}

if (length(merges) == 1){
     # Just do one merge
      ps.menu.id <- merge_taxa(ps.menu.12, merges[[1]])
} else {
     # Merge once to make new phyloseq
     ps.menu.id <- merge_taxa(ps.menu.12, merges[[1]])
     # Then loop over remainder
     for (i in 2:length(merges)){
          ps.menu.id <- merge_taxa(ps.menu.id, merges[[i]])
     }
}

taxa_names(ps.menu.id) <- tax_table(ps.menu.id)[, 'mb_to_mp_to_menu']
ps.menu.id
```
```{r}
# Metabarcoding
tax_table(ps.mb) <- 
     name.sync %>% 
     select(mb_to_mp_to_menu, mb_taxa) %>% 
     right_join(data.frame(ps.mb@tax_table@.Data),
               by = c('mb_taxa' = 'name')) %>%
     distinct() %>% 
     column_to_rownames(var = 'mb_taxa') %>% 
     as.matrix()

ps.mb.id <- ps.mb
taxa_names(ps.mb.id) <- ps.mb.id@tax_table@.Data[, 'mb_to_mp_to_menu']
```

```{r}
# Metaproteomics
# Add lowest level name column
tax_table(ps.protein) <- 
     ps.protein@tax_table@.Data |> 
     data.frame() |> 
     MButils::lowest_level() |> 
     as.matrix()

tax_table(ps.protein) <- 
     name.sync %>% 
     select(mb_to_mp_to_menu, mp_taxa) %>% 
     right_join(data.frame(ps.protein@tax_table@.Data),
               by = c('mp_taxa' = 'name')) %>%
     distinct() %>% 
     column_to_rownames(var = 'mp_taxa') %>% 
     as.matrix()
```

Doing a glom on shared name doesn't work if there's a higher taxonomic level that differs, e.g.
- Musa acuminata / banana
- Musa x paradisiaca / banana 
will both be preserved, when I actually want them to be merged

```{r}
# Get a list of these duplicates
dups <- 
     ps.protein@tax_table@.Data %>% 
     data.frame() %>% 
     filter(duplicated(mb_to_mp_to_menu)) %>% 
     pull(mb_to_mp_to_menu) |> 
     unique()

dups
```

```{r}
# Note this has to be done in a different way if there's >1 duplicate,
# see menu below
ps.protein
merges = list()
for (taxon in dups){
     # Get the list of equivalent taxa at the species level for that food
     eq_taxa <- 
          ps.protein@tax_table@.Data %>% 
          data.frame() %>% 
          filter(mb_to_mp_to_menu == taxon) %>% 
          select(-mb_to_mp_to_menu) %>% 
          MButils::lowest_level() %>% 
          pull(name)
     
     merges <- append(merges, list(eq_taxa))
}

if (length(merges) == 1){
     # Just do one merge
      ps.protein.id <- merge_taxa(ps.protein, merges[[1]])
} else {
     # Merge once to make new phyloseq
     ps.protein.id <- merge_taxa(ps.protein, merges[[1]])
     # Then loop over remainder
     for (i in 2:length(merges)){
          ps.protein.id <- merge_taxa(ps.protein.id, merges[[i]])
     }
}

taxa_names(ps.protein.id) <- tax_table(ps.protein.id)[, 'mb_to_mp_to_menu']
ps.protein.id
```
```{r}
# Now how many taxa are present in each of the objects about to be compared?
cat('Recorded:', ntaxa(ps.menu.id), '\n')
cat('DNA:', ntaxa(ps.mb.id), '\n')
cat('Protein:', ntaxa(ps.protein.id), '\n')
```

# Analyze 

### All taxa

```{r}
# Melt objects and join together
melt.menu <- 
     ps.menu.id |> 
     psmelt() |> 
     select(Sample,
            common_name = OTU,
            abd_menu = Abundance,
            kingdom)

melt.mb <- 
     ps.mb.id |> 
     psmelt() |> 
     select(Sample,
            common_name = OTU,
            abd_dna = Abundance,
            kingdom)

melt.mp <- 
     ps.protein.id |> 
     psmelt() |> 
     select(Sample,
            common_name = OTU,
            abd_protein = Abundance,
            kingdom)
```

```{r}
# Join together
melt <- 
     full_join(melt.menu,
               melt.mb,
               by = c('Sample',
                      'kingdom',
                      'common_name')) |> 
     full_join(melt.mp,
               by = c('Sample',
                      'kingdom',
                      'common_name'))

n_distinct(melt$common_name)
```

```{r}
cat('Expected # taxa after merge is', 
    length(union(union(taxa_names(ps.menu.id), 
                   taxa_names(ps.mb.id)),
             taxa_names(ps.protein.id)))
)
```

Note that this strategy makes entries that weren't detected by a measure NA. For the menu data, we'll convert these to FALSE-- an item was detected by a molecular measure but not recorded in the menu.  Will leave the DNA and protein NAs as-is because I want these to be represented visually as gaps on the plot.

```{r}
melt$abd_menu[is.na(melt$abd_menu)] <- 0
```

```{r}
# Make abundances binary
melt <- 
     mutate(melt,
            across(starts_with('abd_'),
                   function(x){x > 0}))
```

```{r}
# Label predictions 
# If failed samples present (OTU table entries are NAs), need to define a 
# version of "&" that does not return FALSE for NA & FALSE 
and_na <- function(x, y){
     
     # Instead returns NA
     return <- NULL
     if (any(is.na(c(x,y)))){
          # If there's also FALSE, change the output
          if (FALSE %in% c(x, y)){
               return <- NA
          } else {
          # Otherwise keep as-is
               return <- x & y
          }
     } else {
          return <- x & y
     }
     
     return
}
```

```{r}
# Label predictions
# True positives
melt$tp_dna <- mapply(and_na, melt$abd_dna, melt$abd_menu)
melt$tp_protein <- mapply(and_na, melt$abd_protein, melt$abd_menu)

# True negatives
melt$tn_dna <- mapply(and_na, !melt$abd_dna, !melt$abd_menu)
melt$tn_protein <- mapply(and_na, !melt$abd_protein, !melt$abd_menu)

# False positives
melt$fp_dna <- mapply(and_na, melt$abd_dna, !melt$abd_menu)
melt$fp_protein <- mapply(and_na, melt$abd_protein, !melt$abd_menu)

# False negatives
melt$fn_dna <- mapply(and_na, !melt$abd_dna, melt$abd_menu)
melt$fn_protein <- mapply(and_na, !melt$abd_protein, melt$abd_menu)
```

```{r}
head(melt)
```
```{r}
# Pivot this data into long form, grouping by food and counting up predictions across samples (TP, TN, FP, FN)
melt <-
     melt |> 
     # Pull info on study from sample name
     mutate(study = substr(Sample, 
                           start = 1, 
                           stop = 2)) |>
     select(-starts_with('abd')) |> 
     pivot_longer(cols = matches('protein|dna'),
                  names_sep = '_',
                  names_to = c('prediction',
                               'marker')) |>
     group_by(study, common_name, kingdom, marker, prediction) |> 
     # Count the number of TRUEs for each prediction
     summarize(n = sum(value, na.rm = TRUE)) 
```

```{r}
# Factor data for visualization
# Calculate F-measure to be used downstream
# For this ranking, combine results from both cohorts
pred.summary <-
     melt %>% 
     select(-study) %>% 
     group_by(common_name, marker, prediction) %>% 
     summarize(n = sum(n)) %>% 
     pivot_wider(names_from = prediction, values_from = n) %>% 
     mutate(recall = tp/(tp + fn),
            precision = tp/(tp + fp),
            f_measure = (2 * recall * precision)/(recall + precision)) %>% 
     arrange(marker, desc(f_measure), desc(tn), desc(fp))

pred.summary
```

```{r}
# Now that joins complete, add factor levels
# Factor foods by DNA first
food.levels <- 
     pred.summary |> 
     filter(marker == 'protein') |> 
     pull(common_name)

melt <- 
     melt %>% 
     mutate(
          study = factor(study,
                         levels = c('HD', 'ID'),
                         labels = c('Habitual', 'Intervetion')),
          marker = factor(marker,
                          levels = c('protein', 'dna'),
                          labels = c('Protein', 'DNA')),
          common_name = factor(common_name, 
                               levels = food.levels),
          prediction = factor(prediction, levels = c('tp',
                                                     'tn',
                                                     'fp',
                                                     'fn'),
                              labels = c('True positive',
                                         'True negative',
                                         'False positive',
                                         'False negative')),
          kingdom = case_when(kingdom == 'Viridiplantae' ~ 'Plants',
                              kingdom == 'Metazoa' ~ 'Animals',
                              kingdom == 'Fungi' ~ 'Other',
                              is.na(kingdom) ~ 'Other'),
          kingdom = factor(kingdom,
                           levels = c('Plants',
                                      'Animals',
                                      'Other')))
```

```{r}
# Save for comparison to metabarcoding data
# saveRDS(melt,
#         here('data',
#              'processed',
#              'performance-metrics',
#              'DNA and protein v. menu predictions by shared food taxa, 1-2d prior, no intake filter.rds'))
```

#### Plot

```{r}
# Add a "Not detected" category for items detected by only one or the other measure, not both
# If this happens, TP + TN + FP + FN = 0 for the marker
not_detected <- 
     melt |> 
     group_by(kingdom, common_name, marker) |> 
     summarize(count = sum(n)) |> 
     filter(count == 0) |> 
     select(-count) |> 
     mutate(prediction = 'Absent from dataset',
            n = 1)

not_detected

# Append to main dataset
melt <- bind_rows(melt, not_detected)

# Refactor prediction category
melt$prediction <- factor(melt$prediction, 
                              levels = c('True positive',
                                         'True negative',
                                         'False positive',
                                         'False negative',
                                         'Absent from dataset'))
```

```{r}
melt %>%
     ggplot(aes(x = common_name, y = n, fill = prediction)) +
     geom_bar(position = 'fill', stat = 'identity') +
     scale_fill_manual(values = c('#309343','#8AC373', # TP, TN
                                  '#D7191C', '#FDAE61',# FP, FN
                                  'gray90')) + # Absent
     facet_grid(cols = vars(kingdom),
                rows = vars(marker),
                scales = 'free',
                space = 'free') +
     theme_classic() +
     theme(axis.text.x = element_text(angle = 40, hjust = 1),
           axis.text.y = element_blank(), 
           axis.ticks = element_blank(),
           axis.line = element_blank(),
           axis.title = element_blank(),
           legend.position = 'none',
           legend.title = element_blank(), 
           plot.margin = margin(0, 0.5, 0.5, 0.5, 'in'),
           strip.background = element_blank()) 
```

Helpful to look at "zoomed out"
But, if interested in taxon labels, can also break into individual rows

```{r}
# Actually think I might need to just generate small "sub-plots" by facet and stitch together in Illustrator because I'm already using two facet variables to make the above plot
# Add facets to go across multiple panels
# pred.long <- 
#      pred.long |> 
#      arrange(common_name) |> 
#      mutate(facet = sort(rep(1:4, 96))) # Tall: 4 rows
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results',
            'manuscript',
            'supplemental',
            paste0(plotID,
                   '_Metabarcoding and metaproteomic by-food predictions compared to menu.pdf')),
       height = 6, width = 24)
```
