---
title: "Performance scatterplots"
output: html_notebook
---

# Setup

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r libraries, echo=FALSE}
library(here) 
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw() +
               theme(
                    axis.text = element_text(size = 12),
                    axis.title.x = ggtext::element_markdown(size = 14),
                    axis.title.y = ggtext::element_markdown(size = 14),
                    legend.title = element_text(size = 12),
                    strip.text = element_text(size = 12)
               )
)
```

# Read in data

## Menu data phyloseq

```{r}
# Menu data phyloseq
ps.menu <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_menu.rds') %>% 
     readRDS()

ps.menu
```
## Metabarcoding phyloseqs

```{r}
# Metabarcoding phyloseqs
ps.animal <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_12SV5.rds') %>% 
     readRDS()

ps.animal

ps.plant <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_mb_trnL.rds') %>% 
     readRDS()

ps.plant
```
## Metaproteomic phyloseq

```{r}
ps.protein <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          # 'ps_mp_1PUP_taxa.rds') %>% # >=1 protein unique peptide
          'ps_mp_1UP_taxa.rds') %>% # >=1 unique peptide
     readRDS()
          
ps.protein
```

# Pre-process

## Menu

```{r}
# This currently has food group info in taxon name:
ps.menu |> 
     taxa_names() |> 
     head()
```
```{r}
# Group by taxon to omit
ps.menu <- 
     ps.menu %>% 
     tax_glom(taxrank = 'name')

# Update names, which previously included food group
taxa_names(ps.menu) <- tax_table(ps.menu)@.Data[, 'name']

ps.menu |> 
     taxa_names() |> 
     head()
```

## Metabarcoding

```{r}
# The 12SV5 data have a shifted taxonomy, maybe due to assignTaxonomy()?
# Below, kingdom should be superkingdom, phylum should be kingdom.
# Think it's class that's throwing this off, for some should be Mammalia
ps.animal@tax_table@.Data
```
```{r}
# Update accordingly
tax_table(ps.animal) <- 
     tax_table(ps.animal)@.Data %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') %>% 
     mutate(superkingdom = kingdom,
            kingdom = phylum,
            phylum = class) %>% 
     select(-class) %>% 
     select(superkingdom, kingdom, phylum, everything()) %>% 
     column_to_rownames(var = 'row') %>% 
     as.matrix()

tax_table(ps.animal)[1:3, 1:3]
```

```{r}
# Note failed trnL samples and relabel with NAs prior to merge
failed <- 
     which(sample_sums(ps.plant)==0) %>% 
     names()

asvtab.plant <- ps.plant@otu_table@.Data
asvtab.plant[failed, ] <- NA
otu_table(ps.plant) <- otu_table(asvtab.plant,
                                 taxa_are_rows = FALSE)

# Combine into a single object
ps.mb <- merge_phyloseq(ps.plant, ps.animal)
ps.mb

# Remove separated objects
rm(ps.plant, ps.animal)
```
## Metaproteomics

Nothing to do here:

```{r}
taxa_names(ps.protein) |> head()
ps.protein
```
## Pair with 1-2 days prior intake

For this broadest level comparison, let's look only at the two days prior to sampling.

```{r}
# These already exist in the sample data:
sample_variables(ps.mb)[grep(pattern = 'delta',
                             sample_variables(ps.mb))]

sample_variables(ps.protein)[grep(pattern = 'delta',
                             sample_variables(ps.protein))]
```

```{r}
samdf.mb <- 
     sample_data(ps.mb) %>% 
     data.frame() %>% 
     rownames_to_column(var = 'row') 
```

```{r}
# Collapse the ASV table entries from the two days prior to each sample into a single entry.  NOTE: Can't use merge_samples on phyloseq here because some samples make up >1 merge

# Do manually on ASV table
asvtab.menu <- otu_table(ps.menu)@.Data
asvtab.menu.12 <- data.frame()

for (sample in seq(dim(samdf.mb)[1])){
     # Pull menu day -1
     one.before <- asvtab.menu[samdf.mb$delta1[sample], ]
     # Pull menu day -2
     two.before <- asvtab.menu[samdf.mb$delta2[sample], ]
     # Merge and place in new, aggregated OTU table
     asvtab.menu.12 <- rbind(asvtab.menu.12,
                             one.before + two.before)
     # Update food names (only has to be done once)
     if (sample == 1){names(asvtab.menu.12) <- names(one.before + two.before)}
     # Update sample name
     row.names(asvtab.menu.12)[sample] <- samdf.mb$row[sample]
}

# Now rebuild a subsetted phyloseq object
ps.menu.12 <- phyloseq(otu_table(asvtab.menu.12, taxa_are_rows = FALSE),
                       sample_data(ps.mb), # Now this matches
                       tax_table(ps.menu)) # Menu-specific taxonomy

# Remove any taxa that aren't present any longer
ps.menu.12 <- prune_taxa(taxa_sums(ps.menu.12) > 0, ps.menu.12)
ps.menu.12
```
```{r}
# How many taxa are present in each of the objects about to be compared?
cat('Recorded:', ntaxa(ps.menu.12), '\n')
cat('DNA:', ntaxa(ps.mb), '\n')
cat('Protein:', ntaxa(ps.protein), '\n')
```

## Reference taxa by their shared family

```{r}
ps.menu.12 <- tax_glom(ps.menu.12, taxrank = 'family')
ps.protein <- tax_glom(ps.protein, taxrank = 'family')
```

```{r}
# Note that tax_glom doesn't like NA entries (for failed plant entries in metabarcoding dataset)
otu_table(ps.mb)[is.na(otu_table(ps.mb))] <- 0
ps.mb <- tax_glom(ps.mb, taxrank = 'family')
```

```{r}
# Update names to the family level
taxa_names(ps.menu.12) <- tax_table(ps.menu.12)@.Data[, 'family']
taxa_names(ps.mb) <- tax_table(ps.mb)@.Data[, 'family']
taxa_names(ps.protein) <- tax_table(ps.protein)@.Data[, 'family']
```

```{r}
# Now, let's go back and re-index the failed samples so they have NAs for the plant families
plant_families <- 
     ps.mb@tax_table@.Data |> 
     data.frame() |> 
     filter(kingdom == 'Viridiplantae') |> 
     pull(family)

otu_table(ps.mb)@.Data[failed, plant_families] <- NA
```

## Combine

```{r}
# Melt objects and join together
melt.menu <- 
     ps.menu.12 |> 
     psmelt() |> 
     select(Sample,
            common_name = OTU,
            abd_menu = Abundance,
            kingdom)

melt.mb <- 
     ps.mb |> 
     psmelt() |> 
     select(Sample,
            common_name = OTU,
            abd_dna = Abundance,
            kingdom)

melt.mp <- 
     ps.protein |> 
     psmelt() |> 
     select(Sample,
            common_name = OTU,
            abd_protein = Abundance,
            kingdom)
```

```{r}
# Join together
melt <- 
     full_join(melt.menu,
               melt.mb,
               by = c('Sample',
                      'kingdom',
                      'common_name')) |> 
     full_join(melt.mp,
               by = c('Sample',
                      'kingdom',
                      'common_name'))

n_distinct(melt$common_name)
```

```{r}
cat('Expected # taxa after merge is', 
    length(union(union(taxa_names(ps.menu.12), 
                   taxa_names(ps.mb)),
             taxa_names(ps.protein)))
)
```

Note that this strategy makes entries that weren't detected by a measure NA. For the menu data, we'll convert these to FALSE-- an item was detected by a molecular measure but not recorded in the menu.  Will leave the DNA and protein NAs as-is because I want these to be represented visually as gaps on the plot.

```{r}
melt$abd_menu[is.na(melt$abd_menu)] <- 0
```

```{r}
# Make abundances binary
melt <- 
     mutate(melt,
            across(starts_with('abd_'),
                   function(x){x > 0}))
```

```{r}
# Label predictions 
# If failed samples present (OTU table entries are NAs), need to define a 
# version of "&" that does not return FALSE for NA & FALSE 
and_na <- function(x, y){
     
     # Instead returns NA
     return <- NULL
     if (any(is.na(c(x,y)))){
          # If there's also FALSE, change the output
          if (FALSE %in% c(x, y)){
               return <- NA
          } else {
          # Otherwise keep as-is
               return <- x & y
          }
     } else {
          return <- x & y
     }
     
     return
}
```

```{r}
# Label predictions
# True positives
melt$tp_dna <- mapply(and_na, melt$abd_dna, melt$abd_menu)
melt$tp_protein <- mapply(and_na, melt$abd_protein, melt$abd_menu)

# True negatives
melt$tn_dna <- mapply(and_na, !melt$abd_dna, !melt$abd_menu)
melt$tn_protein <- mapply(and_na, !melt$abd_protein, !melt$abd_menu)

# False positives
melt$fp_dna <- mapply(and_na, melt$abd_dna, !melt$abd_menu)
melt$fp_protein <- mapply(and_na, melt$abd_protein, !melt$abd_menu)

# False negatives
melt$fn_dna <- mapply(and_na, !melt$abd_dna, melt$abd_menu)
melt$fn_protein <- mapply(and_na, !melt$abd_protein, melt$abd_menu)
```

```{r}
head(melt)
```
```{r}
# Pivot this data into long form, grouping by food and counting up predictions across samples (TP, TN, FP, FN)
melt <-
     melt |> 
     # Pull info on study from sample name
     mutate(study = substr(Sample, 
                           start = 1, 
                           stop = 2)) |>
     select(-starts_with('abd')) |> 
     pivot_longer(cols = matches('protein|dna'),
                  names_sep = '_',
                  names_to = c('prediction',
                               'marker')) |>
     group_by(study, common_name, kingdom, marker, prediction) |> 
     # Count the number of TRUEs for each prediction
     summarize(n = sum(value, na.rm = TRUE)) 
```

```{r}
# Factor data for visualization
# Calculate F-measure to be used downstream
# For this ranking, combine results from both cohorts
pred.summary <-
     melt %>% 
     select(-study) %>% 
     group_by(common_name, marker, prediction) %>% 
     summarize(n = sum(n)) %>% 
     pivot_wider(names_from = prediction, values_from = n) %>% 
     mutate(recall = tp/(tp + fn),
            precision = tp/(tp + fp),
            f_measure = (2 * recall * precision)/(recall + precision)) %>% 
     arrange(marker, desc(f_measure), desc(tn), desc(fp))

pred.summary
```

```{r}
# Now that joins complete, add factor levels
# Factor foods by DNA first
food.levels <- 
     pred.summary |> 
     filter(marker == 'protein') |> 
     pull(common_name)

melt <- 
     melt %>% 
     mutate(
          study = factor(study,
                         levels = c('HD', 'ID'),
                         labels = c('Habitual', 'Intervetion')),
          marker = factor(marker,
                          levels = c('protein', 'dna'),
                          labels = c('Protein', 'DNA')),
          common_name = factor(common_name, 
                               levels = food.levels),
          prediction = factor(prediction, levels = c('tp',
                                                     'tn',
                                                     'fp',
                                                     'fn'),
                              labels = c('True positive',
                                         'True negative',
                                         'False positive',
                                         'False negative')),
          kingdom = case_when(kingdom == 'Viridiplantae' ~ 'Plants',
                              kingdom == 'Metazoa' ~ 'Animals',
                              kingdom == 'Fungi' ~ 'Other',
                              is.na(kingdom) ~ 'Other'),
          kingdom = factor(kingdom,
                           levels = c('Plants',
                                      'Animals',
                                      'Other')))
```

# Visualize 

```{r}
# Add a "Not detected" category for items detected by only one or the other measure, not both
# If this happens, TP + TN + FP + FN = 0 for the marker
not_detected <- 
     melt |> 
     group_by(kingdom, common_name, marker) |> 
     summarize(count = sum(n)) |> 
     filter(count == 0) |> 
     select(-count) |> 
     mutate(prediction = 'Absent from dataset',
            n = 1)

not_detected

# Append to main dataset
melt <- bind_rows(melt, not_detected)

# Refactor prediction category
melt$prediction <- factor(melt$prediction, 
                              levels = c('True positive',
                                         'True negative',
                                         'False positive',
                                         'False negative',
                                         'Absent from dataset'))
```

```{r}
melt %>%
     ggplot(aes(x = common_name, y = n, fill = prediction)) +
     geom_bar(position = 'fill', stat = 'identity') +
     scale_fill_manual(values = c('#309343','#8AC373', # TP, TN
                                  '#D7191C', '#FDAE61',# FP, FN
                                  'gray90')) + # Absent
     facet_grid(cols = vars(kingdom),
                rows = vars(marker),
                scales = 'free',
                space = 'free') +
     theme_classic() +
     theme(axis.text.x = element_text(angle = 40, hjust = 1),
           axis.text.y = element_blank(), 
           axis.ticks = element_blank(),
           axis.line = element_blank(),
           axis.title = element_blank(),
           legend.position = 'none',
           legend.title = element_blank(), 
           plot.margin = margin(0, 0.5, 0.5, 0.5, 'in'),
           strip.background = element_blank()) 
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")
ggsave(here('results',
            'manuscript',
            '3',
            paste0(plotID,
                   '_Molecular family predictions compared to menu.pdf')),
       height = 4, width = 15)
```

