---
title: "Visualize menu data"
output: html_notebook
---

## Setup

```{r setup}
require(knitr)
knitr::opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(cowplot)
library(ggdendro)
library(here)
library(phyloseq)
library(tidyverse)
```

```{r}
# Plotting themes
theme_set(theme_bw())
          # Some error here interacting w/ggdendrogram below
          #      theme(
          #           axis.text = element_text(size = 12),
          #           axis.title.x = ggtext::element_markdown(size = 14),
          #           axis.title.y = ggtext::element_markdown(size = 14),
          #           legend.title = element_text(size = 12),
          #           strip.text = element_text(size = 12)
          #      )
```

## Read in data

### Menu data

```{r}
# Tax-glommed
ps.protein <- 
     here('data',
          'processed',
          'phyloseq',
          'combined',
          'filtered',
          'ps_mp_1UP.rds') %>%
     readRDS()

ps.protein
```

```{r}
ps.menu <- 
     here('data', 
          'processed', 
          'phyloseq',
          'combined',
          'filtered',
          'ps_menu.rds') %>% 
     readRDS()

ps.menu
```

### Common names

```{r}
name.sync <- 
     here('data',
          'processed',
          'Naming discrepancies.xlsx') %>% 
     readxl::read_excel()

name.sync
```

## Pre-process

### List samples

ML48 (10/17/19) omitted due to low spectra counts and poor extraction.

```{r}
samdf.protein <- data.frame(ps.protein@sam_data)
samdf.protein
```
```{r}
# Add missing sample
row <- 
     data.frame(subj = 'ML48',
                date = as.Date('2019-10-17'),
                study = 'Intervention',
                time = NA
     ) %>% 
     mutate(delta1 = paste(subj, date-1, sep = '_'),
            delta2 = paste(subj, date-2, sep = '_'))

row.names(row) <- paste(row$subj, 
                        row$date,
                        sep = '_')

row
```
```{r}
samdf.protein <- 
     bind_rows(samdf.protein, row) %>% 
     rownames_to_column(var = 'row')
```

### Rename taxa

```{r}
# Menu
tax_table(ps.menu) <- 
     name.sync %>% 
     select(common_names, menu_taxa) %>% 
     right_join(data.frame(ps.menu@tax_table@.Data), 
               by = c('menu_taxa' = 'name')) %>% 
     select(food_group, name = menu_taxa, common_names) %>%
     distinct() %>% 
     mutate(row = paste(
          name,
          food_group,
          sep = '_'
     )) %>% 
     column_to_rownames(var = 'row') %>% 
     as.matrix()
```

### 1-2 days' prior

Maybe 1 day prior to sampling would be better, so foods are not shared between rows? i.e., if samples are from consecutive days, will happen that comparison menu for day 1 [*day 0* and day -1] will overlap with menu for day 2 [day 1 and *day 0*], making them seem more similar.  Here, we care more about similarity between individual menu days.

```{r}
# 1 day prior
ps.menu.subset <- 
     ps.menu %>% 
     prune_samples(samdf.protein$delta1, .) %>% 
     prune_taxa(taxa_sums(.) > 0, .)

ps.menu.subset
```

## Visualize

First pass
Try with food groups only

```{r}
melt <- 
     psmelt(ps.menu.subset)
```

```{r}
ggplot(melt, aes(x = OTU, y = Sample, fill = Abundance>0)) +
     geom_tile() +
     scale_fill_manual(values = c('white', '#7e756d')) +
     theme(axis.text.y = element_blank(),
           axis.ticks.y = element_blank())
```
Consider excluding some food groups to simplify?

```{r}
melt %>% 
     filter(Abundance > 0) %>% 
     group_by(food_group, name) %>% 
     count() %>% 
     group_by(food_group) %>% 
     summarize(n_tot = length(n)) %>% 
     arrange(desc(n_tot))
```
Removing "other" and "oils-sugars" will drop 29 entries (think this is reasonable-- could still use them in the clustering?)

```{r}
melt %>% 
     filter(!(food_group %in% c('other',
                                'oils-sugars'))) %>% 
     ggplot(aes(x = OTU, y = Sample, fill = Abundance>0)) +
     geom_tile() +
     scale_fill_manual(values = c('white', '#7e756d')) +
     theme(axis.text.y = element_blank(),
           axis.ticks.y = element_blank())
```
A smidge more manageable-- try moving forward with this. 
Update phyloseq object used for clustering to reflect this change:

```{r}
ntaxa(ps.menu.subset)
ps.menu.subset <- 
     ps.menu.subset %>% 
     subset_taxa(food_group != 'other') %>% 
     subset_taxa(food_group != 'oils-sugars')
ntaxa(ps.menu.subset)

# Remove these from melted object too
melt <- filter(melt, !(food_group %in% c('other',
                                         'oils-sugars')))
```

## Add dendrogram

### Samples
```{r}
# Run clustering
menu.dat <- as.matrix(ps.menu.subset@otu_table@.Data)
# Make data binary:
menu.dat[menu.dat > 0] <- 1
menu.dendro <- 
     vegan::vegdist(menu.dat,
                    method = 'jaccard') %>% 
     hclust(method = 'average') %>% 
     as.dendrogram()

# Create dendro
dendro.plot.y <- ggdendrogram(data = menu.dendro, rotate = TRUE)

# Preview the plot
dendro.plot.y
```

```{r}
# Re-order samples to match
menu.order <- order.dendrogram(menu.dendro)
menu.order

melt <- 
     mutate(melt,
            Sample = factor(Sample,
                            levels = row.names(menu.dat[menu.order, ])))
```

### Foods 

```{r}
# Read in labels from dendrogram file
ifsac.labs <- 
     here('data',
          'processed', 
          'IFSAC dendrogram.csv') %>% 
     read_csv() %>% 
     pull(food_group)

head(ifsac.labs)
```


```{r}
# Transcribe IFSAC groupings into an hclust object
ifsac <- list()  # initialize empty object
# define merging pattern: 
#    negative numbers are leaves, 
#    positive are merged clusters (defined by row number in $merge)
ifsac$merge <- matrix(c(-2, -3,   # 1
                        -4, -5,   # 2
                        -7, -8,   # 3
                        -1, 1,    # 4
                        2, -6,    # 5
                        5, 3,     # 6
                        4, 6,     # 7
                        -10, -11, # 8
                        -13, -14, # 9
                        -9, 8,    # 10
                        -12, 9,   # 11
                        11, -15,  # 12
                        10, 12,   # 13
                        -17, -18, # 14
                        -19, -20, # 15
                        -22, -23, # 16
                        14, 15,   # 17
                        -21, 17,  # 18
                        16, 18,   # 19
                        -24, -25, # 20
                        -26, -27, # 21
                        20, 21, # 22
                        19, 22, # 23
                        -16, 23, # 24
                        13, 24, # 25
                        7, 25 # 26
                        ), nc = 2, byrow = TRUE) 
ifsac$height <- c(rep(4, 2),
                  rep(5, 3),
                  6, 7,
                  rep(2, 2),
                  rep(3, 2),
                  4, 5, 
                  rep(1, 3),
                  2, 3, 4,
                  rep(3, 2),
                  4, 5, 6, 7, 8) # define merge heights
ifsac$order <- 1:27      # leaf order (trivial if hand-entered)
ifsac$labels <- ifsac.labs    # labels of leaves
class(ifsac) <- "hclust"        # make it an hclust object
plot(ifsac)                     # look at the result   

#convert to a dendrogram object if needed
ifsac.d <- as.dendrogram(ifsac)
```
```{r}
# # Run clustering
# menu.dat <- as.matrix(ps.menu.subset@otu_table@.Data)
# # Make data binary:
# menu.dat[menu.dat > 0] <- 1
```

```{r}
menu.plot <- 
     ggplot(menu.dendro.data$segments) + 
     geom_segment(aes(x = x, y = y, xend = xend, yend = yend)) +
     geom_tile(data = menu.dendro.data$labels,
               aes(x, y, fill = kingdom),
               height = 0.2) +
     scale_fill_manual(values = c('#b66353',
                                  '#d7ce9f',
                                  '#638b66')) 
  # geom_text(data = menu.dendro.data$labels, aes(x, y, label = label),
  #           hjust = 1, angle = 90, size = 3) 
```

```{r}
# Re-order samples to match
menu.order <- order.dendrogram(menu.dendro)
menu.order

melt <- 
     mutate(melt,
            OTU = factor(OTU,
                         levels = colnames(menu.dat[, rev(menu.order)])))
```

## Joint plot

```{r}
# Now, update original heatmap with these new levels
heatmap <- 
     ggplot(melt, aes(x = Sample, y = OTU, fill = Abundance>0)) +
     geom_tile() +
     scale_fill_manual(values = c('white', '#7e756d')) +
     coord_equal() +
     theme(axis.text = element_blank(),
           axis.ticks = element_blank(),
           legend.position = 'none') 

heatmap
```

```{r}
plotID <- paste(Sys.Date(), as.character(sample(000:999,1)), sep=".")

ggsave(here('results', 
            'manuscript', 
            '1',
            paste0(plotID, '_Menu visualization.pdf')),
       height = 8)
```

```{r}
plot_grid(menu.plot,
          heatmap,
          aligh = 'h')
```

