---
title: "Diet data to share"
output: html_notebook
---

```{r setup, include=FALSE, echo=FALSE}
require("knitr")
opts_knit$set(root.dir = file.path(getwd(), '..'))
```

```{r}
library(here)
library(tidyverse)
library(phyloseq)
```

# Read in data

## Menu

```{r}
ps.menu <- 
     here('data',
          'processed',
          'phyloseq',
          'combined',
          'filtered',
          'ps_menu.rds') %>% 
     readRDS()

ps.menu
```

## Proteins

```{r}
proteins.df <- 
     here('data',
          'processed',
          'share',
          '5FDR_1UP_all proteins.rds') %>% 
     readRDS()

proteins.df
```
# Pre-process

### Samples

```{r}
# Pull this from metaproteomic data
samples <- 
     proteins.df %>% 
     select(study, subj, date) %>% 
     distinct()

samples
```
```{r}
# Add delta for join to menu
samples <- 
     bind_rows(samples, 
               samples) %>% 
     arrange(study, subj, date) %>% 
     mutate(delta = rep(1:2, 26)) %>% 
     mutate(delta_date = date - delta)
```

### Menu

Subset to taxa of interest: rice and eggs

```{r}
# What's available?
colnames(ps.menu@otu_table@.Data)[grep(colnames(ps.menu@otu_table@.Data), pattern = 'Oryza')]
colnames(ps.menu@otu_table@.Data)[grep(colnames(ps.menu@otu_table@.Data), pattern = 'Gallus')]
```

"Oryza sativa_other" contains ingredients like rice wine vinegar.  Let's subset to grains and eggs for now, and then can add more details on eggs if Alfredo wants yolk/white difference.

```{r}
menu <- 
     ps.menu %>% 
     prune_taxa(c("Oryza sativa_grains",
                  "Gallus gallus_eggs, shell eggs"), .) %>% 
     psmelt() %>% 
     rename(food = OTU) %>% 
     mutate(food = ifelse(food == 'Gallus gallus_eggs, shell eggs',
                          yes = 'eggs',
                          no = 'rice')) %>% 
     select(study, subj, date, food, Abundance) %>% 
     # Recode subject IDs to original
     mutate(subj = factor(subj,
                          levels = c('HD', 
                                     'ID1', 
                                     'ID2',
                                     'ID3',
                                     'ID4'),
                          labels = c('908',
                                     'AK65',
                                     'JN55',
                                     'ML48',
                                     'PC74'))) %>% 
     mutate(subj = as.character(subj))

menu
```
```{r}
# Join to sample data
samples <- 
     samples %>% 
     left_join(menu,
               by = c('study', 
                      'subj', 
                      'delta_date' = 'date'))
```

```{r}
# Pivot wider to have days side-by-side
samples <- 
     samples %>% 
     pivot_wider(names_from = 'food',
                 values_from = 'Abundance') %>% 
     mutate(delta = -delta) %>% 
     select(-delta_date) %>% 
     pivot_wider(names_from = delta,
                 values_from = c(rice, eggs))
```

```{r}
samples
```

## Proteins

Already filtered to only proteins with >1UP and master proteins of the protein group. 

### Duplicate names

```{r}
proteins.df %>% 
     filter(grepl('\n', Accession)) %>% 
     select(Accession, Description, `# PSMs`)
```
```{r}
# Are any of these chicken or rice?
proteins.df %>% 
     filter(grepl('\n', Accession)) %>% 
     filter(grepl('Gall_', Accession)) %>% 
     # select(Accession, Description) %>% 
     # distinct()
     group_by(subj, date) %>%
     summarize(psms_tot = sum(`# PSMs`)) %>%
     pull(psms_tot) 
```
```{r}
proteins.df %>% 
     filter(grepl('\n', Accession)) %>% 
     filter(grepl('Osa_', Accession)) %>% 
     select(Accession, Description, `# PSMs`)
```
Note these to Alfredo and see how I should follow up.

```{r}
write_csv(samples,
          here('data', 
               'processed', 
               'share',
               'Rice and egg intake by sample.csv'))
```

```{r}
# Also write protein data as CSV
write_csv(proteins.df,
          here('data',
               'processed',
               'share',
               '5FDR_1UP_all proteins.csv'))
```

